---
title: "Alpha_diversity_prep"
author: "Stephanie Wilson"
date: "February 2024"
output: html_document
---

## SUMMARY
This script calculates alpha diversity for polyphenol intake, merges it with microbial metagenomic data, and creates a new metadata file.

## INPUTS
  - **FL100_polyphenol_compound_intake.RData** - polyphenol compound intake (n = 350)
  - **species_313.Rdata** - abundance + taxonomy data (n = 313)
  - **metaphlan_faith_pd.csv** - FL100 phylogenetic diversity from Andrew

## OUTPUTS
  - **FL100_metagenome_Meta.RData** - metadata + metagenomic and polyphenol alpha diversity metrics

## PREP DATA  
Load Packages
```{r message=FALSE, warning=FALSE}
library(tidyverse); library(bestNormalize); library(devtools); library(vegan); library(effects); library(ggpubr); library(broom); library(jtools)
```

Load data
```{r}
# polyphenol compound intake data
load(file = "FL100_diet_prep/output/FL100_polyphenol_compound_intake.RData")

shotgun_faith = read.csv("data/metaphlan_faith_pd.csv") %>%
  rename("UserName" = "subject_id")

#metadata files
load(file = "RData/species_313.RData")
load(file = "RData/metadata_313.RData")
```


## Calculation of Compound Alpha Diversity
```{r} 
ppalpha = compound_long %>%
  group_by(UserName) %>%
  summarize(pp_sobs = vegan::specnumber(compound_mg1000kcal), # number of unique polyphenols
            pp_shannon = vegan::diversity(compound_mg1000kcal, index = "shannon"),
            pp_simpson = vegan::diversity(compound_mg1000kcal, index = "simpson"),
            pp_invsimpson = 1/pp_simpson,
            pp_pielous = pp_shannon/log(pp_sobs),
            pp_daily_mg1000kcal = sum (compound_mg1000kcal)) 
```


## Load metagenomic species data and metadata
```{r}
metadata_313 = cleaned_data
metagenome = as.data.frame(t(abundance_313))
  metadata_313$mb_sobs = specnumber(metagenome)
  metadata_313$mb_shannon = vegan::diversity(metagenome, index = "shannon")
  metadata_313$mb_evenness = metadata_313$mb_shannon/(log(metadata_313$mb_sobs))
```

Merge dataframes
```{r}
merge_313 = metadata_313 %>%
  left_join(ppalpha) %>%
  left_join(shotgun_faith)

save(merge_313, file = 'RData/FL100_metagenome_Meta.RData')
```

## Exploratory
```{r}
ppalpha %>%
  pivot_longer(cols = c(pp_sobs, pp_shannon, pp_simpson, pp_invsimpson), names_to = "metric") %>%
  ggplot(aes(x = pp_daily_mg1000kcal, y = value)) + 
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~ metric, ncol =4, scales = "free_y")
```

Polyphenol Diversity Histograms
```{r}
par(mfrow=c(2,2))
hist(merge$pp_sobs)
hist(merge$pp_simpson)
hist(merge$pp_invsimpson)
```

