---
title: "LPS_microbes_prep"
author: "Stephanie Wilson"
date: "December 2024"
output: html_document
---

## SUMMARY
This script aggregates the abundance of specific bacterial species to address questions related to LPS.
  - Bacteroidales: known for being the main producer of LPS in a healthy gut
  - Enterobacteriaceae and Gammaproteobacteria: known for having potent pro-inflammatory LPS

## INPUT
  - **phyloseq_313.Rdata** - phyloseq object for n = 313
  
## OUTPUT
  - **LPS_related_bacteria_metagenomic.csv** - Abundances of select bacterial taxa relevant for LPS Binding Protein Analysis


## Load packages and data  
```{r Load packages}
library(tidyverse); library(phyloseq)
```

```{r Load data}
load('RData/phyloseq_313.RData')
```

## Prepare Class
```{r}
phyloseq_313_class = tax_glom(phyloseq_313, "Class")

gammaproteobacteria_tax = as.data.frame(tax_table(phyloseq_313_class)) %>%
  filter(Class == 'c__Gammaproteobacteria') %>%
  rownames_to_column("otu")

gammaproteobacteria_otu = as.data.frame(otu_table(phyloseq_313_class)) %>%
  rownames_to_column("otu") %>%
  filter(otu %in% gammaproteobacteria_tax$otu)

gammaproteobacteria_sum = gammaproteobacteria_otu %>%
  select(-otu) %>%    # Exclude the 'otu' column
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("UserName") %>%
  rename(Gammaproteobacteria = V1)
```


## Prepare Order
```{r}
phyloseq_313_order = tax_glom(phyloseq_313, "Order")

bacteroidales_tax = as.data.frame(tax_table(phyloseq_313_order)) %>%
  filter(Order == 'o__Bacteroidales') %>%
  rownames_to_column("otu")

bacteroidales_otu = as.data.frame(otu_table(phyloseq_313_order)) %>%
  rownames_to_column("otu") %>%
  filter(otu %in% bacteroidales_tax$otu)

bacteroidales_sum = bacteroidales_otu %>%
  select(-otu) %>%    # Exclude the 'otu' column
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("UserName") %>%
  rename(Bacteroidales = V1)
```

## Prepare Family
```{r}
phyloseq_313_family = tax_glom(phyloseq_313, "Family")

enterobacteriaceae_tax = as.data.frame(tax_table(phyloseq_313_family)) %>%
  filter(Family == 'f__Enterobacteriaceae') %>%
  rownames_to_column("otu")

enterobacteriaceae_otu = as.data.frame(otu_table(phyloseq_313_family)) %>%
  rownames_to_column("otu") %>%
  filter(otu %in% enterobacteriaceae_tax$otu)

enterobacteriaceae_sum = enterobacteriaceae_otu %>%
  select(-otu) %>%    # Exclude the 'otu' column
  summarise(across(everything(), \(x) sum(x, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("UserName") %>%
  rename(Enterobacteriaceae = V1)
```

## Merge Bacterial Abundances
```{r}
merge = gammaproteobacteria_sum %>%
  left_join(bacteroidales_sum) %>%
  left_join(enterobacteriaceae_sum)

write.csv(merge, 'data/LPS_related_bacteria_metagenomic.csv', row.names = FALSE)
```





