---
title: "Polyphenol Class Prep"
author: "Stephanie Wilson"
date: "September 2024"
output: html_document
---

## SUMMARY
This script finds other polyphenol compounds within our list of 3063 polyphenols that fall into polyphenolic classes listed in the dbPUP substrate file.

## INPUT
  - **polyphenol_substrates_long_FooDB_taxonomy.csv**, PUP substrates from dbPUP linked to their matching FooDB ID and taxonomy
  - **PhenolExplorer_linkedto_FooDB.csv**, Phenol Explorer Polyphenols previously linked to their matching FooDB ID
  - **FooDB_phenols_taxonomy.csv**, 3063 polyphenols in FooDB and their taxonomy

## OUTPUT
  - **FooDB_dbPUP_class_list.csv** - dbPUP substrate classes and all underlying substrates and polyphenols from our original set of 3063 polyphenols 

## LOAD PACKAGES AND DATA
packages
```{r}
library(tidyverse)
```

Data Files
```{r}
# FooDB linked dbPUP polyphenol substrates
substrates = read.csv('dbPUP/polyphenol_substrates_long_FooDB_taxonomy.csv')

# Phenol Explorer polyphenols linked to their FooDB equivalent
PE_FooDB_links = read.csv('data/PhenolExplorer_linkedto_FooDB.csv')

# FooDB polyphenol taxonomy
foodb = read.csv('data/FooDB_phenols_taxonomy.csv')
```

Combine substrate list with existing polyphenol list
```{r message = FALSE}
full_list = substrates %>%
  distinct(compound_public_id, .keep_all = TRUE) %>%
  select(c(compound_public_id:directparent)) %>%
  mutate(addition = "new") %>%
  full_join(foodb) %>%
  filter(!is.na(compound_public_id)) %>%
  mutate(addition = ifelse(is.na(addition), 'old', addition))
```

### Identify Compound Classes and identify a search strategy 

The subclasses we need to match
```{r}
substrates %>% distinct(subclass) %>% pull()
```

Search Strategy
1) Pull compounds for each class in dbPUP substrate 
2) Pull compounds for each class in Phenol Explorer-FooDB Matched File
3) Pull compounds using a taxonomy level search in FooDB

dbPUP subclass, FooDB Taxonomy Level
  - Isoflavonoids as klass
  - Flavones as subklass
  - Chalcones and dihydrochalcones combined as subklass
  - Flavanols (aka flavan-3-ols) is there as direct parent
  - Flavanones not there, considered a flavan subkclass
  - Flavonols exist as direct parent
  - Stilbenes, exists as klass
  - Curcuminoids, only one compound
  

### Perform Search
```{r}
isoflavonoids = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Isoflavonoids'] | 
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Isoflavonoids'] |
         grepl(klass, pattern ='Isoflavonoids', ignore.case = TRUE) & 
         !grepl('Homoisoflavonoids', klass, ignore.case = TRUE)) %>%
  mutate(class = 'Isoflavonoids') 

flavones = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Flavones'] |
      compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Flavones'] |
      grepl(subklass, pattern = 'Flavones', ignore.case = TRUE)) %>%
  mutate(class = 'Flavones')
  
chalcones_dihydrochalcones = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Chalcones']|
           compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass 
                                                                     %in% c('Chalcones', "Dihydrochalcones")] |
           grepl(subklass, pattern ='Chalcones', ignore.case = TRUE)) %>%
  mutate(class = 'Chalcones_Dihydrochalcones')

flavanols = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Flavanols']|
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Flavonols'] |
           grepl(directparent, pattern ='flavan-3-ols', ignore.case = TRUE)) %>%
  mutate(class = 'Flavanols_aka_flavan3ols')

flavanones = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Flavanones']|
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Flavanones'] |
         grepl('Flavanones', directparent, ignore.case = TRUE) & 
         !grepl('isoflavanones', directparent, ignore.case = TRUE)) %>%
  mutate(class = 'Flavanones')

flavonols = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Flavonols']|
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Flavonols'] |
           grepl(directparent, pattern ='flavonols', ignore.case = TRUE)) %>%
  mutate(class = 'Flavonols')

stilbenes = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Stilbenes']|
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Stilbenes'] |
           grepl(klass, pattern ='stilbenes', ignore.case = TRUE)) %>%
  mutate(class = 'Stilbenes')

curcuminoids = full_list %>%
  filter(compound_public_id %in% substrates$compound_public_id[substrates$subclass == 'Curcuminoids']|
         compound_public_id %in% PE_FooDB_links$compound_public_id[PE_FooDB_links$subklass == 'Curcuminoids'] |
           grepl(directparent, pattern = 'curcuminoids', ignore.case = TRUE)) %>%
  mutate(class = 'Curcuminoids')
```

Merge search hit dataframes
```{r}
taxonomy_search = bind_rows(isoflavonoids, flavones, flavanones, chalcones_dihydrochalcones,
                            flavanols, flavonols, stilbenes)

# Check for any duplicates
duplicate_keep = taxonomy_search %>%
  group_by(compound_public_id) %>%  # Group to identify duplicate rows
  filter(n() > 1) %>%
  filter(class == 'Flavonols')

taxonomy_search_cleaned = taxonomy_search %>%
  # Remove duplicates
  filter(!compound_public_id %in% duplicate_keep$compound_public_id) %>%
  # Add final decision on duplicates
  full_join(duplicate_keep)
```

Write File
```{r}
write.csv(taxonomy_search_cleaned, 'data/FooDB_dbPUP_class_list.csv', row.names = FALSE)
```


