---
title: "A2b_Substrates_FooDB_Taxonomy"
author: "Stephanie Wilson"
date: "September 2024"
output: html_document
---

## SUMMARY
This script merges the taxonomy of FooDB polyphenols into the PUP substrates file. Web scraping is performed for those polyphenols not already in our FooDB polyphenol taxonomy list (ie, phenolic compounds).

## INPUT
  - **FooDB_phenols_taxonomy.csv**, Previously web scraped taxonomy for our 3063 polyphenols
  - **polyphenol_substrates_long.csv**, PUP substrates from dbPUP
  - **dbPUP_substrates_FooDB_IDs.csv**, PUP substrates from dbPUP linked to their matching FooDB ID
  
## OUTPUT
  - **polyphenol_substrates_long_FooDB_taxonomy.csv**, PUP substrates from dbPUP linked to their matching FooDB ID and taxonomy
  - **polyphenol_substrates_need_comp_data.csv**, PUP substrates from dbPUP linked to their matching FooDB ID and taxonomy, **but there is no linked intake data**
 
## LOAD PACKAGES AND DATA  
Load packages
```{r}
library(tidyverse); library(rvest); library(xml2); library(stringr)
source('functions/get_taxonomy.R')
```

Load Data
```{r}
substrates = read.csv('dbPUP/polyphenol_substrates_long.csv')
substrates_FooDBmatch = read.csv('data/dbPUP_substrates_FooDB_IDs.csv') %>%
  rename(compound_public_id = compound_code)
tax = read.csv('data/FooDB_phenols_taxonomy.csv')
```

Add taxonomy to FooDB matches
```{r}
FooDB_matches_wtax = left_join(substrates_FooDBmatch, tax, by = "compound_public_id") %>%
  # Remove those that were not detected in FooDB
  filter(!decision == "no")
```

### WEB SCRAPE
Some substrates had a match in FooDB but were *not* in our original polyphenol list of 3063 polyphenols. Let's isolate those and build the link.
```{r}
# Sme FoOD Compounds that need taxonomy searched
tax_search_list = FooDB_matches_wtax %>%
  filter(is.na(compound_name)) %>%
  select(-c(compound_name:directparent)) %>%
  mutate(compound_urls = paste0("https://foodb.ca/compounds/", compound_public_id))

# We will need to pull food composition data from these compounds, so write the file
write.csv(tax_search_list, 'data/polyphenol_substrates_need_comp_data.csv', row.names = FALSE)
```

Run taxonomy scrape function on list
```{r}
# Initialize a list to store the results
taxonomy_results = list()

# Loop through each URL in the compound_urls column
for (i in 1:length(tax_search_list$compound_urls)) {
  
  # Apply the get_taxonomy function to each URL
  result = get_taxonomy(tax_search_list$compound_urls[i])
  
  # Store the result in the list
  taxonomy_results[[i]] = result
}

# Convert to a dataframe
taxonomy_df = do.call(rbind, taxonomy_results) 
taxonomy_hits = cbind(tax_search_list, taxonomy_df) %>%
  select(-compound_urls)
```

### MERGE FILES
Clean up our FooDB matches file
```{r}
FooDB_matches_final = FooDB_matches_wtax %>%
  #Take out what we just searched 
  filter(!compound_public_id %in% tax_search_list$compound_public_id) %>%
  # Add new hits back in
  full_join(taxonomy_hits)
```

 Combine our matches file with the substrates_long file 
```{r}
substrates_wtax = substrates %>%
  left_join(FooDB_matches_final, by = "polyphenol", relationship ="many-to-many")
```

Save this output
```{r}
write.csv(substrates_wtax, 'dbPUP/polyphenol_substrates_long_FooDB_taxonomy.csv', row.names = FALSE)
```

