---
title: "A2d_Substrate_ASA24_Match"
author: "Stephanie Wilson"
date: "September 2024"
output: html_document
---

## SUMMARY
This script takes content of new polyphenols (not in the original 3063) and maps this to FL100 dietary data. Resulting output is an RDS file that has polyphenol intake summarized at different levels (unaggregated, by recall, by subject).

# INPUT

-   **FL100_diet/ingredientized_asa_10-2022.csv** - Ingredientized FL100 ASA24
-   **codematched_final.csv**: FL100 ASA to FooDB mapping key
-   **FooDB_Content_Extracted_dbPUP_substrates.csv**\*

# OUTPUT

-   **FL100_FooDB_dbPUP_Substrate_Intakes.RDS**, 3 dataframes of FL100 polyphenol intakes (with new polyphenols) from:
    1)  finest level unaggregated
    2)  by recall
    3)  by subject

## LOAD PACKAGES AND DATA
Load packages
```{r}
library(tidyverse); library(readxl)
```

Load Data
```{r}
content = read.csv('data/FooDB_Content_Extracted_dbPUP_substrates.csv') %>%
  relocate(orig_content_avg_RFadj, .after =  orig_content_avg) %>%
  rename(food_id = food_public_id)

matched = read.csv('data/FL100_diet/codematched_final.csv') %>%
  relocate('food_V2_ID', 'orig_food_common_name', 
           .before = Ingredient_code) %>%
  select(-Ingredient_description)

#Ingredientized ASA
ingredients = read.csv('data/FL100_diet/ingredientized_asa_10-2022.csv') %>%
  select(-c(WWEIA.Category.description, UserID, EatWith, KCAL:D_CHEESE, 
            Seq.num, PortionCode, Lang, RecallStatus))
```

### 1) Merge dietary recall data with FooDB matches

```{r}
ASA = left_join(ingredients, matched, by = 'Ingredient_code') %>%
  relocate(Ingredient_code, Ingredient_description, food_id, food_V2_ID, 
           orig_food_common_name, mean_score, .after = Food_Description) %>%
  relocate(RecallNo, .after = UserName) %>%
  select(-c(food_V2_ID, orig_food_id, fndds_description, 
            IntakeStartDateTime, IntakeEndDateTime, RecallAttempt))

## Merge FooDB-matched ASA Ingredient Codes with FooDB Polyphenol Content File.
## Link between FooDB Polyphenol Content and ASA code-matched data is *food_id*.
ASA_hits = ASA %>%
  dplyr::left_join(content, by = 'food_id', relationship = 'many-to-many') %>%
  # keep foods that only occur in our content file
  filter(food_id %in% content$food_id)
```

```{r}
# What do we have quantified
ASA_hits_quant = ASA_hits %>%
  filter(!is.na(orig_content_avg))

# What is predicted but not quantified.
ASA_hits_unquant = ASA_hits %>%
  filter(is.na(orig_content_avg))
```

```{r}
ASA_by_compound = ASA_hits_quant %>%
  dplyr::mutate(
    pp_consumed = case_when(
      
      # Whey Shakes (excluding chocolate whey shakes)
      FoodCode == 95230000 ~ (orig_content_avg_RFadj * 0.01) * (Ingredient_consumed_g * 0.02),
      
      # Nectar Beverages
      FoodCode == 64203020 ~ (orig_content_avg_RFadj * 0.01) * (Ingredient_consumed_g * 0.15),
      FoodCode == 64204010 ~ (orig_content_avg_RFadj * 0.01) * (Ingredient_consumed_g * 0.20),
      
      # Tea Adjustment (only for 'DUKE' citation)
      food_id %in% c(1021, 957, 38, 940, 849, 939, 517, 748) & citation == 'DUKE' ~ 
        (orig_content_avg_RFadj * 0.01) * (Ingredient_consumed_g * (1/50)),

      # Coffee Adjustment (only for 'DUKE' citation)
      food_id %in% c(891, 60, 58, 59) & citation == 'DUKE' ~ 
        (orig_content_avg_RFadj * 0.01) * (Ingredient_consumed_g * (10/180)),

      # Default case: Standard calculation for all other cases
      TRUE ~ (orig_content_avg_RFadj * 0.01) * Ingredient_consumed_g
    )
  )
```

### 2) Reformat into summary files

```{r}
ASA_by_recall = ASA_by_compound %>%
  
  # Sum by Recall and Participant
  group_by(UserName, RecallNo) %>%
  mutate(pp_recallsum = sum(pp_consumed, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(UserName, RecallNo, .keep_all = TRUE) 

ASA_by_subject = ASA_by_recall %>%
  
  # Average by Participant
  group_by(UserName) %>%
  mutate(pp_average_mg = mean(pp_recallsum, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(UserName, .keep_all = TRUE) %>%
  select(-c(RecallNo, pp_recallsum))
```

Save Data

```{r}
save(ASA_by_compound, ASA_by_recall, ASA_by_subject, file = 'RData/FL100_FooDB_dbPUP_Substrate_Intakes.RDS')
```
