---
title: "A2e Quantify FL100 Intakes at Class and Compound level"
author: "Stephanie Wilson"
date: "September 2024"
output: html_document
---

## SUMMARY
This script takes FL100 polyphenol intake data and specifically aggregates intake at the class and compound level by participant.

## INPUT
  - **FooDB_dbPUP_class_list.csv** - dbPUP substrate classes and all underlying substrates and polyphenols from our original set of 3063 polyphenols 
  - **FL100_PP_Raw.csv** - FL100 Polyphenol Intakes of original 3063 set, by compound, food, recall, and subject
  - **FL100_merged_variables.csv** - FL100 metadata variables
  - **FL100_FooDB_dbPUP_Substrate_Intakes.RDS**, 3 dataframes showing FL100 polyphenol intakes of new polyphenols from 1) finest level unaggregated to 2) by recall, then 3) by subject
  
## OUTPUT
  - **FL100_dbPUPsubstrate_class_intakes.csv** - FL100 intakes for dbPUP substrate classes, including broader class groupings using our original set of 3063 polyphenols, adjusted by caloric intake
  - **FL100_dbPUPsubstrate_compound_intakes.csv** - FL100 polyphenol substrate intakes by compound, adjusted by caloric intake

## LOAD packages and data
Load packages
```{r}
library(tidyverse)
```

Load Data Files
```{r}
# Compounds and their classes
class_list = read.csv('data/FooDB_dbPUP_class_list.csv') %>%
  select(c(compound_public_id, class)) 

# Load in Dietary Polyphenol Content, 
# Raw is Each compound for each food in each recall by participant
PP_Raw = read.csv('data/FL100_PP_Raw.csv') %>%
  # REmove compound outliers
  filter(!compound_name %in% c('Chinese tannin', 'alpha-Viniferin'))

## New Substrate hits matched to ASA24
load('RData/FL100_FooDB_dbPUP_Substrate_Intakes.RDS')

# Load partcipant metadata for kcal
kcal = read.csv('data/FL100_merged_variables.csv') %>%
  select(c(UserName, avg_total_kcal))
```

## Combine substrate and PP_Raw Lists then Filter in relevant compounds
```{r message = FALSE}
# Remove extra columns
ASA_by_recall_coltrim = ASA_by_recall %>%
  select(intersect(names(ASA_by_recall), names(PP_Raw)))

# Merge Files
merged = full_join(PP_Raw, ASA_by_recall_coltrim) %>%
  # There are two compounds that fit the bill for Rutin
  filter(compound_public_id %in% c(class_list$compound_public_id, 'FDB002536'))%>%
  # Let's Add those classes in
  left_join(class_list, by = 'compound_public_id')
```

## Calculate Averge Polyphenol Class Intakes
```{r}
class_intakes = merged  %>%
  
  #Group by Taxonomic Class
  group_by(UserName, RecallNo, class) %>%
  
  #gets the sum of each compound for each participant's recall
  mutate(class_intake_mg = sum(pp_consumed, na.rm = TRUE)) %>% 
  select(c(UserName, RecallNo, class, class_intake_mg)) %>%
  ungroup()%>%
  
  #Remove duplicates since we've summed each polyphenol per recall
  distinct(UserName, RecallNo, class, .keep_all = TRUE) %>%
  
  #Average polyphenol intake across recalls for each class
  group_by(UserName, class) %>%
  mutate(Avg_class_intake = mean(class_intake_mg)) %>%
  ungroup() %>%
  
  #Remove duplicates
  distinct(UserName, class, .keep_all = TRUE) %>%
  select(c(UserName, class, Avg_class_intake)) %>%
  filter(!is.na(class)) %>%
  
  # Add kcal data
  left_join(kcal, by = 'UserName') %>%
  mutate(class_intake_mg1000kcal = Avg_class_intake/(avg_total_kcal/1000)) %>%
  
  # Remove kcal and avg_class intake now that we have intake standardized to kcal
  select(-c(Avg_class_intake, avg_total_kcal)) %>%
  
  #Transpose dataframe where each column is a participant
  pivot_wider(id_cols = UserName, names_from = class, 
              values_from = class_intake_mg1000kcal, values_fill = 0)
```

```{r}
write.csv(class_intakes, 'output/FL100_dbPUPsubstrate_class_intakes.csv', row.names = FALSE)
```

## Calculate Averge Polyphenol Compound Intakes
```{r}
compound_intakes = merged  %>%
  
  #Group by Compound ID
  group_by(UserName, RecallNo, compound_public_id) %>%
  
  #gets the sum of each compound for each participant's recall
  mutate(compound_intake_mg = sum(pp_consumed, na.rm = TRUE)) %>% 
  select(c(UserName, RecallNo, compound_public_id, compound_name, compound_intake_mg)) %>%
  ungroup()%>%
  
  #Remove duplicates since we've summed each polyphenol per recall
  distinct(UserName, RecallNo, compound_public_id, .keep_all = TRUE) %>%
  
  #Average polyphenol intake across recalls for each class
  group_by(UserName, compound_public_id) %>%
  mutate(Avg_compound_intake = mean(compound_intake_mg)) %>%
  ungroup() %>%
  
  #Remove duplicates
  distinct(UserName, compound_public_id, .keep_all = TRUE) %>%
  select(c(UserName, compound_public_id, compound_name, Avg_compound_intake)) %>%
  filter(!is.na(compound_public_id))  %>% 
  # Add kcal data
  left_join(kcal, by = 'UserName') %>%
  mutate(compound_intake_mg1000kcal = Avg_compound_intake/(avg_total_kcal/1000)) %>%
  
  # Remove kcal and avg_class intake now that we have intake standardized to kcal
  select(-c(Avg_compound_intake, avg_total_kcal)) %>%
  
  #Transpose dataframe where each column is a participant
  pivot_wider(id_cols = UserName, names_from = compound_public_id, 
              values_from = compound_intake_mg1000kcal, values_fill = 0)
```

```{r}
write.csv(compound_intakes, 'output/FL100_dbPUPsubstrate_compound_intakes.csv', row.names = FALSE)
```
