---
title: "PUP diversity and ML preparation"
author: "Stephanie Wilson"
date: "July 2025"
output: html_document
---

## SUMMARY
This script calculates PUP-diversity and formats our microbial data for taxaHFE usage. Regarding microbial data specifically, this:
- **microbial PUP gene count**: formats PUP data into hierarchical data (class > family > protein gene) for taxaHFE usage
- **abundance PUP-containing microbial genera**: formats microbial abundance data into hierarchical data (taxonomy-based) for taxaHFE usage

## INPUT
-   **FL100_metagenome_species_full_names.csv**, microbial abundance data from MetaPhlan
-   **metadata_313.Rdata**, study covariates, polyphenol class intakes, GI markers
-   **PUP_taxonomy_modelling_data.Rdata**
    -   meta: FL100 data including age, BMI, sex, fiber intake, total HEI score, and total polyphenol intake
    -   PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
    -   tax: microbial PUP count data in long form
    -   gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
    -   class_filtered: class intakes that were in at least 30% of our cohort
    -   compound_filtered: compound intakes that were in at least 30% of our cohort dbpup_overview_clean.csv - PUP gene classifications metadata_313.Rdata - includes covariates for LBP analysis
-   **dbpup_overview_clean.csv**, classification and details of microbial PUP genes

## OUTPUT

-   PUP_alpha_diversity_metrics.csv 
-   PUP_matrix.csv
-   PUP_meta_LBP.csv
-   PUP_MB_matrix.csv

## PREP

```{r packages}
library(tidyverse); library(vegan)
```

```{r data, message = FALSE}
# PUP counts
load("RData/PUP_taxonomy_modelling_data.Rdata")

# Abundance, From A0b_microbiome_prep
species_tax = read.csv('data/FL100_metagenome_species_full_names.csv')
colnames(species_tax) = ifelse(grepl("^X\\d", colnames(species_tax)),
                       sub("^X", "", colnames(species_tax)),
                       colnames(species_tax))

# metadata and LBP
load("RData/metadata_313.Rdata")
metadata = cleaned_data %>%
  select(c(UserName, Age, Sex, BMI, fiber_g1000kcal, ppintake_mg1000kcal, 
           hei_asa24_totalscore, plasma_lbp_bd1))%>%
  mutate(ppintake_quartile = ntile(ppintake_mg1000kcal, 4))

## Load dbpup classifications
dbpup = read.csv('dbPUP/dbpup_overview_clean.csv') %>%
  select(c(class, 'Family.Number', gene))
```

## Calculate PUP alpha-diversity metrics

```{r}
# Step 1: Aggregate to counts per subject and PUP gene
subject_pup = tax %>%
  rename(UserName = subject_id) %>%
  filter(!is.na(genus)) %>%
  group_by(UserName, gene) %>%
  summarise(reads = sum(read_count), .groups = "drop")

# Step 2: Create a wide table with genus as columns
matrix = subject_pup %>%
  pivot_wider(names_from = gene, values_from = reads, values_fill = 0) %>%
  column_to_rownames("UserName")  # vegan expects rownames

# Step 3: Calculate richness and evenness
richness = rowSums(matrix > 0)
shannon = diversity(matrix, index = "shannon")
pielou_evenness = shannon / log(richness)

# Step 4: Combine into one data frame
diversity_metrics = tibble(
  subject_id = rownames(matrix),
  richness = richness,
  shannon = shannon,
  pielou_evenness = pielou_evenness)

write.csv(diversity_metrics, 'data/PUP_alpha_diversity_metrics.csv', row.names = FALSE)
```

## Prepare PUP files for taxa HFE

```{r}
matrix_hfe = matrix %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  left_join(dbpup) %>%
  relocate(class, Family.Number, gene, .before = '5001') %>%
  mutate(gene = paste("g__", gene, sep = ""),
         class = paste("c__", class, sep = ""),
         Family.Number = paste("f__", Family.Number, sep = "")) %>%
  unite("clade_name", class:gene, sep = "|", remove = FALSE) %>%
  select(-c(class:gene)) 

meta_hfe = metadata %>%
  select(c(UserName, Age, Sex, BMI, fiber_g1000kcal, hei_asa24_totalscore, plasma_lbp_bd1)) %>%
  rename(Sample = UserName)

# CSVs
write.csv(matrix_hfe, 'taxaHFE_PUP/PUP_matrix.csv', row.names = FALSE)
write.csv(meta_hfe, 'taxaHFE_PUP/PUP_meta_LBP.csv', row.names = FALSE)
```

## Prepare microbial files for taxaHFE

Filter microbial genus data for only those that mapped to PUP genes. Then format into a hierachical format for taxaHFE usage.

```{r}
# Pull out genus names 
genus_list = gene_genus_combo %>%
  distinct(genus) %>%
  mutate(search_term = str_remove(genus, "^g__"))

# create the search pattern to pull out genera associated with PUP
pattern = str_c(genus_list$search_term, collapse = "|")

# GENUS LEVEL
genus_input = species_tax %>%
  # Pull in genera that our genes mapped to
  filter(str_detect(Genus, pattern)) %>%
  # Collapse the taxonomic names
  unite("clade_name", Kingdom:Genus, sep = "|", remove = FALSE) %>%
  # Keep only relevant content, and subjects
  select(clade_name, all_of(as.character(meta$subject_id))) %>%
  # Ensure everything is summarized by genus
  group_by(clade_name) %>%
  summarise_all(.funs = sum, na.rm = TRUE)

# CSVs
write.csv(genus_input, 'taxaHFE_PUP_MB/PUP_MB_matrix.csv', row.names = FALSE)
write.csv(meta_hfe, 'taxaHFE_PUP_MB/PUP_meta_LBP.csv', row.names = FALSE)
```
