---
title: "D1_PUP_alpha_diversity"
author: "Stephanie Wilson"
date: "July 2025"
output: html_document
---

## SUMMARY
This script performs linear models to assess whether: 
1) polyphenol intake alpha diversity predicts PUP gene alpha diversity
2) PUP gene alpha diversity predicts LBP 

## INPUT
-   **PUP_alpha_diversity_metrics.csv** - microbial and polyphenol intake alpha diversity metrics
-   **FL100_metagenome_Meta.RData**

## OUTPUT
-   **PUP_alpha_diversity_models.csv** - output from six PUP alpha diversity models, including models predicting LBP

## Prep
```{r packages, message = FALSE}
library(tidyverse); library(car); library(ggResidpanel); library(broom)
```

```{r data}
# PUP gene diversity
pup_diversity = read.csv('data/PUP_alpha_diversity_metrics.csv') %>%
  rename(UserName = subject_id,
         pup_sobs = richness,
         pup_pielou = pielou_evenness,
         pup_shannon = shannon)

# FL100 metadata including GI markers and polyphenol intake diversity metrics.
load('Rdata/FL100_metagenome_Meta.RData')

# Merge dataframes
diversity = merge_313 %>%
  select(c(UserName, Age, Sex, BMI, fiber_g1000kcal, 
           hei_asa24_totalscore, plasma_lbp_bd1,
           mb_sobs:pp_pielous)) %>%
  left_join(pup_diversity, by = "UserName")
```

## Predicting PUP richness
```{r}
pup_sobs_richness_lm = lm(pup_sobs ~ pp_sobs + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
summary(pup_sobs_richness_lm)

pup_sobs_evenness_lm = lm(pup_sobs ~ pp_pielous + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
summary(pup_sobs_evenness_lm)
```

```{r}
resid_panel(pup_sobs_richness_lm)
resid_panel(pup_sobs_evenness_lm)
```

## Predicting PUP evenness
```{r}
pup_evenness_richness_lm = lm(pup_pielou ~ pp_sobs + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
summary(pup_evenness_richness_lm)

pup_evenness_evenness_lm = lm(pup_pielou ~ pp_pielous + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
summary(pup_evenness_evenness_lm)
```

```{r}
resid_panel(pup_evenness_richness_lm)
resid_panel(pup_evenness_evenness_lm)
```

## Predicting LBP
```{r}
lbp_pup_richness_lm = lm(log(plasma_lbp_bd1) ~ pup_sobs + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)

lbp_pup_evenness_lm = lm(log(plasma_lbp_bd1) ~ pup_pielou + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
```

```{r}
lbp_mb_richness_lm = lm(log(plasma_lbp_bd1) ~ mb_sobs + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)

lbp_mb_evenness_lm = lm(log(plasma_lbp_bd1) ~ mb_evenness + Age + Sex+ BMI + fiber_g1000kcal + hei_asa24_totalscore, data = diversity)
```

## Combine Model Outputs
```{r}
models = list(
  pup_richness_intake_richness = pup_sobs_richness_lm,
  pup_richness_intake_evenness = pup_sobs_evenness_lm,
  pup_evenness_intake_richness = pup_evenness_richness_lm,
  pup_evenness_intake_evenness = pup_evenness_evenness_lm,
  lbp_pup_richness = lbp_pup_richness_lm,
  lbp_pip_evenness = lbp_pup_evenness_lm,
  lbp_mb_richness = lbp_mb_richness_lm,
  lbp_mb_evenness = lbp_mb_evenness_lm
)

combined_models = map_dfr(models, tidy, .id = "model")

write.csv(combined_models, 'output/PUP_alpha_diversity_models.csv', row.names = FALSE)
```

## Examine significant relationships
```{r}
sig_variables = combined_models %>%
  filter(p.value < 0.05)
```

