---
title: "Converting_taxaHFE_ML_files_for_dietML"
author: "Stephanie Wilson"
date: "2025-07-24"
output: html_document
---

## Summary
This script takes taxaHFE's ML files (test_data.csv, train_data.csv) and performs a full join on rows into a singular file. Then it splits by column so to generate metadata.csv and matrix.csv.

## Input
Repeated for PUP model and PUP-containing microbial abundance model.
- test_data.csv
- train_data.csv

## Output
Repeated for PUP model and PUP-containing microbial abundance model.
- metadata.csv
- matrix.csv

## Prep
```{r message = FALSE}
library(tidyverse)
```

## Create custom function.
```{r}
convert_taxaHFEoutput_for_ML = function(dir, output_dir = "dietML") {

  # Pull out files given a base directory  
  train_path = file.path(dir, "train_data.csv")
  test_path  = file.path(dir, "test_data.csv")
  
  # Read files in
  train = read.csv(train_path, check.names = FALSE)
  test  = read.csv(test_path, check.names = FALSE)
  
  # Combine on rows
  full_data = full_join(train, test) %>%
    arrange(subject_id)
  
  # Meta data columns of interest
  metadata_cols = c("age", "sex", "bmi", "fiber_g1000kcal", "hei_asa24_totalscore")
  
  # Split by Columns - metadata
  df_metadata = full_data %>% select(subject_id, all_of(metadata_cols))
  
  # Split by Columns - features
  df_features = full_data %>%
    select(subject_id, everything()) %>%
    select(-any_of(setdiff(metadata_cols, "subject_id")))
  
  # Write converted outputs
  write.csv(df_metadata, file.path(dir, output_dir, "df_metadata.csv"), 
            row.names = FALSE)
  write.csv(df_features, file.path(dir, output_dir, "df_matrix.csv"), 
            row.names = FALSE)
  
  list(metadata = df_metadata, features = df_features)
}
```

## Apply Function to directories of interest
dietML directy must exist in these first.
```{r}
# Define base directories
dirs = c("taxaHFE_PUP", "taxaHFE_PUP_MB")

# Apply function
results = lapply(dirs, convert_taxaHFEoutput_for_ML)
```

