---
title: "SHAP Plot for best PUP Gene Random Forest Model"
author: "Stephanie Wilson"
date: "July 2025"
output: html_document
---

## SUMMARY
This script creates a publication ready SHAP plot for the best performing ML model predicting LBP with microbial PUP genes.

## INPUT
- **dbpup_overview_clean.csv** - dbPUP gene classificiations
- **ML_r_workspace.rds** - packaged results from random forest model. 

## OUTPUT
- **ML_LBP_PUP_SHAP.png**
- **ML_LBP_PUP_SHAP.rds**, image components to be combined with the other best model.

## DATA PREP
```{r}
library(tidyverse)
```

```{r}
# Load PUP information
dbpup = read.csv('dbPUP/dbpup_overview_clean.csv')
```

## Establish key for formatting SHAP plot feature names
```{r}
covariates = tibble(
  feature = c("age", "sex_male", "bmi", "fiber_g1000kcal", "hei_asa24_totalscore"),
  feature_publish = c("Age", "Sex, Male", "BMI", "Fiber (g/1000 kcal)", "Total HEI Score"))

## Load dbpup classifications
feature_list = dbpup %>%
  select(class, Family.Number, gene, Name) %>%
  pivot_longer(cols = c(class, Family.Number, gene),
               names_to = "variable",
               values_to = "value") %>%
  mutate(Name = if_else(variable == "gene", Name, NA_character_),
         value = if_else(variable == "class",
                         str_to_sentence(str_replace_all(value, "_", " ")),
                         value),
         # Format feature names with UniProt Number
         feature_publish = case_when(
            variable == "gene" ~ paste(Name, " (", value, ")", sep = ""),
            variable == "class" ~ paste(value, sep = ""),
            variable == "Family.Number" ~ paste(value, sep = "")),
         # Allow long feature names to go on multiple lines
         feature_publish = ifelse(
            str_length(feature_publish) > 20,
            str_wrap(feature_publish, width = 20),
            feature_publish),
         # Add in formatting to pair with what is present in dietML data
         feature = case_when(
            variable == "gene" ~ paste("g_", str_to_lower(value), sep = ""),
            variable == "class" ~ paste("c_", str_to_lower(value), sep = ""),
            variable == "Family.Number" ~ paste("f_", str_to_lower(value), sep = ""))) %>%
  select(feature, feature_publish) %>%
  full_join(covariates) 
```

## Create SHAP Plot
```{r}
# Path to best performing seed
bestmodel = "taxaHFE_PUP/dietML/output/ml_results_seed_527/ML_r_workspace.rds"  
loaded_env = load(bestmodel)
  
  # Extract the data with SHAP values, match to formatted names
  sv_data = data.frame(feature = colnames(sv_full[[2]])) %>%
    left_join(feature_list, by = 'feature') %>%
    distinct(feature, feature_publish)
  
  # Update the column names with formatted names
  colnames(sv_full$X) = sv_data$feature_publish
  colnames(sv_full$S) = sv_data$feature_publish
  
  # Create the beeswarm plot
  pup_plot = shapviz::sv_importance(sv_full, kind = "bee", show_numbers = TRUE,
                                bee_width = 0.2,  max_display = 5, 
                                show_other = FALSE) + 
    theme_bw(base_size = 11) + 
    labs(x = '') + 
    theme(plot.margin = margin(0, 0, 0, 0, "cm"))
```

Export Plot
```{r}
saveRDS(pup_plot, "images/ML_LBP_PUP_SHAP.rds") # For combining with other ML plot
ggsave(plot = pup_plot, filename = "images/ML_LBP_PUP_SHAP.png", 
       units = "in", height = 6, width = 8, dpi = 300)
```


