---
title: "Performance Metrics for PUP-MB Random Forest Models"
author: "Stephanie Wilson"
date: "July 2025"
output: html_document
---

## SUMMARY
This script pools ML output from across dietML output RDS files, which contain numerous dataframes relevant to the model including performance and SHAP values. For models on PUP-containing microbes, this script:

1) Isolates the pathways for each of the RDS files created by one run of dietML
2) Extracts performance metrics from each random forest model

## INPUT
- *dietML RDS files* - from dietML output
  
## OUTPUT
- *ML_LBP_PUP_MB_performance_metrics.csv*
- *ML_LBP_PUP_MB_Seed_Performance.png* - Performance across seeds
- *ML_LBP_PUP_MB_Avg_Performance.png* - Change from Null for best seed

## PREP
```{r LOAD PACKAGES, message=FALSE}
library(tidyverse)
```

Load Functions
```{r}
source('functions/find_rds.R')
source('functions/extract_performance_metrics.R')
```

### 1) Extract Path Information

Get the paths of all rds files
```{r}
rds_files = find_rds_files('taxaHFE_PUP_MB/')
rds_info = tibble(
  path = rds_files,
  # Extract seed number
  seed = str_extract_all(path, "\\d+"))%>%
  rename(path = 1,
         seed = 2) %>%
  slice(-1)
```

### 2) Performance metrics
```{r warning = FALSE}
all_metrics = map_dfr(rds_info$path, ~extract_performance(.x, rds_info)) %>%
  pivot_longer(
    cols = c(estimate, estimate_null),
    names_to = "model",
    values_to = "estimate"
  ) %>%
  mutate(
    model = if_else(model == "estimate", "test", "null")
  )

#Export
write.csv(all_metrics, 'output/ML_LBP_PUP_MB_performance_metrics.csv', row.names = FALSE)
```


## PLOT PERFORMANCE METRICS
```{r}
delta_metrics = all_metrics %>%
  pivot_wider(names_from = model, values_from = estimate) %>%
  filter(!is.na(test), !is.na(null), null != 0) %>%  # avoid divide by zero
  mutate(
    percent_change = (test - null) / null * 100)

summary_delta = delta_metrics %>%
  group_by(metric) %>%
  summarise(
    mean_change = mean(percent_change, na.rm = TRUE),
    ci_low = t.test(percent_change)$conf.int[1],
    ci_high = t.test(percent_change)$conf.int[2],
    .groups = "drop")
```

### By seed
```{r}
seed_plot = ggplot(delta_metrics, aes(x = factor(seed), y = percent_change)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Percent Change (Test vs Null)",
    x = "Seed",
    y = "Percent Change (%)",
    color = "Metric") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

seed_plot
ggsave(plot = seed_plot, filename = 'images/ML_LBP_PUP_MB_Seed_Performance.png', units = "in", dpi =300, height = 4, width = 6.5)
```

Seed 527 was the best performing over the null model.

### Average and 95% CI
```{r}
ci_plot = ggplot(summary_delta, aes(x = metric, y = mean_change)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    x = "Metric",
    y = "Change from Null (%)") +
  theme_bw()

saveRDS(ci_plot, "images/ML_LBP_PUP_MB_Avg_Performance.rds")
ggsave(plot = ci_plot, filename = 'images/ML_LBP_PUP_MB_Avg_Performance.png', units = "in", dpi =300, height = 4, width = 6.5)
```


 
