---
title: "SHAP Plot for Best PUP-containing microbial Random Forest Model"
author: "Stephanie Wilson"
date: "July 2025"
output: html_document
---

## SUMMARY
This script creates a publication ready SHAP plot for the best performing ML model predicting LBP with microbial PUP genes.

## INPUT
- dbpup_overview_clean.csv - dbPUP gene classifications
- ML_r_workspace.rds - packaged results from random forest model. 

## OUTPUT
- **ML_LBP_PUP_MB_SHAP.rds**
- **ML_LBP_PUP_MB_SHAP.png**, image components to be combined with the other best model.

## DATA PREP
```{r}
library(tidyverse)
```

## Establish covariate key for formatting SHAP plot feature names
```{r}
covariates = tibble(
  feature = c("age", "sex_male", "bmi", "fiber_g1000kcal", "hei_asa24_totalscore"),
  feature_publish = c("Age", "Sex, Male", "BMI", "Fiber (g/1000 kcal)", "Total HEI Score"))
```

## Create SHAP Plot
```{r}
# Path to best performing seed
bestmodel = "taxaHFE_PUP_MB/output_seed_527/ML_r_workspace.rds"  
loaded_env = load(bestmodel)

# Extract the data with SHAP values, match to formatted names
  sv_data = data.frame(feature = colnames(sv_full[[2]])) %>%
  left_join(covariates) %>%
  mutate(
    feature = as.character(feature),
    feature_publish = if_else(
      feature_publish == "" | is.na(feature_publish),
      feature %>%
        str_replace("^[^_]*_", "") %>%              # remove before first underscore
        str_replace_all("_", " ") %>%               # replace underscores with spaces
        str_to_sentence(),                          # capitalize first word
      feature_publish))

  # Update the column names with formatted names
  colnames(sv_full$X) = sv_data$feature_publish
  colnames(sv_full$S) = sv_data$feature_publish
  
  # Create the beeswarm plot
  pup_plot = shapviz::sv_importance(sv_full, kind = "bee", show_numbers = TRUE,
                                bee_width = 0.2,  max_display = 5, 
                                show_other = FALSE) + 
    theme_bw(base_size = 11) + 
    labs(x = '') + 
    theme(plot.margin = margin(0, 0, 0, 0, "cm"))
```

Export plot
```{r}
saveRDS(pup_plot, "images/ML_LBP_PUP_MB_SHAP.rds") # for combining with other ML plot
ggsave(plot = pup_plot, filename = "images/ML_LBP_PUP_MB_SHAP.png", 
       units = "in", height = 6, width = 8, dpi = 300)
```


