---
title: "ML_Correlations"
author: "Stephanie Wilson"
date: "August 2025"
output: html_document
---

# SUMMARY
This script plots the relationships between top ML model features and LBP.

## INPUT
  - **df_matrix.csv** - Input into PUP ML model
  - **full_data.csv** - Input into PUP-MB ML model
  - **ML_r_workspace.rds** - 2 files representing ML outputs from seed 527

## OUTPUT
  - **ML_correlations.png** - supplementary figure to go with ML SHAP plots

## DATA PREP  
Load packages
```{r}
library(tidyverse)
```

Load Data
```{r}
pup_feautures = read.csv("taxaHFE_PUP/dietML/df_matrix.csv")
mb_features = read.csv("taxaHFE_PUP_MB/full_data.csv")

# Merge Features
ml_features = left_join(pup_feautures, mb_features)
```

## Pull out original data from top 5 features from PUP and PUP-MB models
We will use these to filter our data.
```{r message = FALSE, warning = FALSE}
load("taxaHFE_PUP_MB/output_seed_527/ML_r_workspace.rds")
mb_features_527 = data.frame(feature = colnames(sv_full[[2]]))%>%
  # Pull out top 5, based on shap plots
  slice(1,5,6,10,11)%>%
  pull(feature)

load("taxaHFE_PUP/dietML/output/ml_results_seed_527/ML_r_workspace.rds")
pup_features_527 = data.frame(feature = colnames(sv_full[[2]])) %>%
   # Pull out top 5, based on shap plots
  slice(2, 3, 5, 13, 14) %>%
  pull(feature)
```

Subset data
```{r}
pup_top5 = ml_features %>%
  select(c(feature_of_interest, "sex", any_of(pup_features_527)))

pup_mb_top5 = ml_features %>%
  select(c(feature_of_interest, "sex", any_of(mb_features_527)))

top5_features_both = full_join(pup_mb_top5, pup_top5)
```

## Spearman correlations with LBP
```{r}
# Format to long
long_data = top5_features_both %>%
  select(-sex) %>%  # exclude sex
  pivot_longer(cols = -feature_of_interest, names_to = "variable", values_to = "value")

# Calculate correlations and create label with correct positioning.
cor_labels = long_data %>%
  group_by(variable) %>%
  summarise(
    test = list(cor.test(value, feature_of_interest, method = "spearman")),
    y_pos = max(feature_of_interest, na.rm = TRUE) * 1.1,
    .groups = "drop") %>%
  mutate(
    spearman_rho = map_dbl(test, ~ signif(.x$estimate, 2)),
    p_value = map_dbl(test, ~ .x$p.value),
    p_label = ifelse(p_value < 0.001, "< 0.001", signif(p_value, 2)),
    label = paste0("ρ = ", spearman_rho, ", p = ", p_label)) %>%
  select(variable, y_pos, label)
```

## Scatterplots - Continuous Variables
Create more informative labels
```{r}
facet_labels = c(
  bmi = "BMI~(kg/m^2)",
  fiber_g1000kcal = "Fiber~(g/1000~kcal)",
  hei_asa24_totalscore = "Total~HEI~Score",
  g_a1s0b1 = "Glycoside~hydrolase~(A1S0B1)~count",
  g_ruthenibacterium = "italic('Ruthenibacterium')~Abundance",
  g_lachnospiraceae_unclassified = "italic('Lachnospiraceae')~'unc.'~Abundance")
```

Create Scatterplots
```{r}
correlation_plots= ggplot(long_data, aes(x = value, y = feature_of_interest)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~variable, scales = "free_x", 
             labeller = labeller(variable = facet_labels, .default = label_parsed)) +
  geom_text(
    data = cor_labels,
    aes(x = -Inf, y = y_pos, label = label),
    hjust = -0.1,
    inherit.aes = FALSE,
    size = 3) +
  labs(
    x = NULL,
    y = "Plasma LBP (µg/mL)") +
  theme_bw() +
  theme(
    axis.title.y = element_text(angle = 90, hjust =0))
```

## Barplot - Factor Variables
```{r}
factor_plot = ggplot(pup_top5, aes(x = sex, y = feature_of_interest)) +
  geom_violin(scale ="width", fill = "gray80") +
  geom_jitter(width = 0.1, alpha = 0.6, color = "black") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
  stat_compare_means(method = "wilcox.test", 
                     label.y = max(pup_top5$feature_of_interest, na.rm = TRUE)*1.1,
                     label.x = "Female",
                     size =3) +
  ylim(0, 45) +
  labs(x = NULL, y = NULL) +
  theme_bw()
factor_plot
```

## Combine Plots
```{r}
# Create a narrower version of factor_plot by combining with spacers
narrow_factor_plot =  factor_plot + plot_spacer() +plot_spacer()+
  plot_layout(ncol = 3, widths = c(1, 1, 1))

combined_plot = correlation_plots / narrow_factor_plot + plot_layout(heights = c(2.5, 1))
combined_plot

ggsave(combined_plot, file = 'images/ML_correlations.png', dpi = 300, width = 7.5, height = 6.5, units = "in")
```

