---
title: "PUP_Taxonomy_Modelling_Prep"
author: "Stephanie Wilson"
date: "March 2025"
output: html_document
---

## SUMMARY
Prepare dataframes from modelling the relationship between polyphenol intake and microbial abundance/microbial PUP count and packages them into an Rdata file.

## INPUT
- metadata
- polyphenol intake data at the class and compound level
- PUP summary data 
- PUP names 
- tax, microbial count data

## OUTPUT
- **PUP_taxonomy_modelling_data.Rdata** - contains PUPnames, meta, tax, gene_genus_combo, class_filtered, compound_filtered

## DATA PREP
Load Packages
```{r message = FALSE}
library(tidyverse); library(dplyr, exclude = "select"); library(Rfit); library(ggtext)
```

Load Data
```{r}
# Pull in metadata for modelling covariates
load("RData/metadata_313.Rdata")

meta = cleaned_data %>%
  select(c(UserName, Age, Sex, BMI, fiber_g1000kcal,
           hei_asa24_totalscore, ppintake_mg1000kcal)) %>%
  rename(subject_id = UserName) 
```

Polyphenol Intakes
```{r}
class_intakes = read.csv('output/FL100_dbPUPsubstrate_class_intakes.csv')
compound_intakes = read.csv('output/FL100_dbPUPsubstrate_compound_intakes.csv') 
```

## Clean PUP gene names for merging into our microbial PUP gene count file.
```{r}
# GENE COUNTS & GENE NAMES
PUPnames = read.csv('dbPUP/dbpup_overview_clean.csv') %>%
  select(c(gene, Name, class)) %>%
  rename(PUPclass = class,
         PUPname = Name) %>%
  mutate(PUPclass = str_replace_all(PUPclass, "_", " "))

# MICROBIAL TAXONOMY FOR PUP HITS
tax = read_tsv(gzfile("taxa_all/full_taxonomic_results_counts.tsv.gz"), show_col_types = FALSE) %>%
  rename(gene = info_2) %>%
  left_join(PUPnames) %>%
  relocate(PUPname, PUPclass, .after = gene) %>%
  # Keep just bacteria and participants with metagenome and diet data
  filter(kingdom == "k__Bacteria") %>%
  filter(subject_id %in% meta$subject_id)
```

## Filter in classes that were in at least 30% of our participant's diets
```{r}
class_keep = class_intakes %>%
  filter(UserName %in% meta$subject_id) %>%
  select(-UserName) %>%
  summarise(across(everything(), ~ sum(. == 0))) %>%
  pivot_longer(cols = everything(), names_to = "class", values_to = "zero_count") %>%
  filter(zero_count <= 0.7*nrow(meta))

class_filtered = class_intakes %>%
  select(any_of(c("UserName", class_keep$class)))
```
No classes were removed.

## Filter in compounds that were in at least 30% of our participant diets
```{r}
compounds_keep = compound_intakes %>%
  filter(UserName %in% meta$subject_id) %>%
  select(-UserName) %>%
  summarise(across(everything(), ~ sum(. == 0))) %>%
  pivot_longer(cols = everything(), names_to = "compound", values_to = "zero_count") %>%
  filter(zero_count <= 0.7*nrow(meta))

compound_filtered = compound_intakes %>%
  select(any_of(c("UserName", compounds_keep$compound)))
```
Initial Compound numbers: 191
Final Compound numbers: 29

## Create Dataframe with counts then create dataframe with unique combos of genes and genera
```{r}
tax_genus= tax %>%
  filter(!is.na(genus)) %>%
  filter(!genus=="g__NA") %>%
  select(c(gene, PUPname, genus, species, subject_id, read_count)) %>%
  # Add metadata
  left_join(meta)

# all unique combinations of PUPgene and genus
gene_genus_combo = tax_genus %>%
  group_by(gene, PUPname, genus) %>%
  mutate(user_count = n_distinct(subject_id)) %>%
  ungroup() %>%
  # FILTER those that are not prevalent in a minimum 10% of participants.
  filter(user_count > 313*0.1) %>%
  distinct(gene, PUPname, genus, .keep_all = TRUE) %>%
  select(c(gene, PUPname, genus, user_count))
```

## Save all of this for modelling!
```{r}
save(PUPnames, meta, tax, gene_genus_combo, class_filtered, compound_filtered, 
     file = "RData/PUP_taxonomy_modelling_data.Rdata")
```
