---
title: "PUP_Correlation_PUPcount"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

# SUMMARY
Assessing the impact of polyphenol intake (total, class, compound) on microbial count using Spearman Correlation. 

## INPUT
PUP_taxonomy_modelling_data.Rdata
  - meta: FL100 participant data including age, BMI, sex, total fiber intake, total HEI score, and total polyphenol intake
  - PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
  - tax: microbial PUP count data in long form
  - gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
  - class_filtered: class intakes that were in at least 30% of our cohort
  - compound_filtered: compound intakes that were in at least 30% of our cohort

## OUTPUT
  - **PUP_count_correlation_totalintake.csv**
  - **PUP_count_correlation_classintake.csv**
  - **PUP_count_correlation_compoundintake.csv**

## DATA PREP
Load packages
```{r}
library(tidyverse)
```

Load Data
```{r}
load("RData/PUP_taxonomy_modelling_data.Rdata")
```

## Filter our microbial abundance data by the pattern
```{r}
filtered_genus_long = tax %>%
  # Pull in gene and genus combos that are in a minimum 30% of participants
  filter(gene %in% gene_genus_combo$gene) %>%
  filter(!str_detect(genus, "unclassified")) %>%
  filter(!is.na(genus)) %>%
  # Get read count sum for each unique instance of gene and genus for each subject
  group_by(gene, PUPname, genus, subject_id) %>%
  mutate(read_count_sum = sum(read_count)) %>%
  distinct(gene, PUPname, genus, subject_id, .keep_all = TRUE) %>%
  ungroup() %>%
  select(-read_count) %>%
  # Merge participant data
  filter(subject_id %in% meta$subject_id) %>%
  left_join(meta)
```

## Total Modelling
For each unique gene and genus combination, run a rank-based estimation regression assessing the impact of total polyphenol intake on microbial PUP count
```{r}
# Create empty lists to store results and skipped combinations
total_list = list()

for (i in seq_len(nrow(gene_genus_combo))) {
  
  gene_i = gene_genus_combo$gene[i]
  pup_i = gene_genus_combo$PUPname[i]
  genus_i = gene_genus_combo$genus[i]
  
  message("Running model ", i, ": gene = ", gene_i, ", PUPname = ", pup_i, ", genus = ", genus_i)
  
  # Prepare the data with the unique combinations
  subset = filtered_genus_long %>%
    filter(gene == gene_i & PUPname == pup_i & genus == genus_i) 
  
  # Perform Spearman correlation
    cor_test = cor.test(subset$read_count_sum, subset$ppintake_mg1000kcal,
                        method = "spearman", exact = FALSE)
    
    # Format output
    output = tibble(
      gene = gene_i,
      PUPname = pup_i,
      genus = genus_i,
      Variable = "ppintake_mg1000kcal",
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
  
  # Add output to a list
  total_list[[length(total_list) + 1]] = output
}
```

Combine and format model output
```{r}
# Combine results from list into a singular table
total_results = bind_rows(total_list)

# Filter poor models before multiple correction adjustment
total_results_with_corrections = total_results %>%
  # Remove models that had too many ties and were unable to form accurate estimates 
  # Indicated by variables showing up as numeric entries 1, 2, ...7
  filter(!str_detect(Variable, "^\\d+$")) %>%
  group_by(gene) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(total_results_with_corrections, 'output/PUP_count_correlation_totalintake.csv', row.names = FALSE)
```


## Class Modelling
```{r}
filtered_genus_long_class = filtered_genus_long %>%
  left_join(class_filtered, by = c("subject_id" = "UserName"))
```

Loop
```{r}
# Create empty lists to store results and skipped combinations
class_results_list = list()

class_list = colnames(class_filtered)[2:length(class_filtered)]

# Run model for each polyphenol class
for (class_i in class_list) {
  message("Running models for class: ", class_i)
  
  # Then for each class, subset the data for each unique instance of gene and genus
  # to analyze the impact of polyphenol intake
  for (i in seq_len(nrow(gene_genus_combo))) {
    
    gene_i = gene_genus_combo$gene[i]
    pup_i = gene_genus_combo$PUPname[i]
    genus_i = gene_genus_combo$genus[i]
  
  message("Running model ", i, ": gene = ", gene_i, ", PUPname = ", pup_i, ", genus = ", genus_i)

    # Subset the data for each unique instance of PUP and genus
    subset = filtered_genus_long_class %>%
    filter(gene == gene_i & PUPname == pup_i & genus == genus_i)
    
    # Pull in class data
    select_class = subset[[class_i]]
    
     # Perform Spearman correlation
    cor_test = cor.test(subset$read_count_sum, select_class,
                        method = "spearman", exact = FALSE)
    
    output = tibble(
      gene = gene_i,
      PUPname = pup_i,
      genus = genus_i,
      class = class_i, 
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
    
    class_results_list[[length(class_results_list) + 1]] <- output
  }
}
```

Combine and format model output
```{r}
# Combine results from list into a singular table
class_results = bind_rows(class_results_list)

# Filter poor models before multiple correction adjustment
class_results_with_corrections = class_results %>%
  group_by(gene) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(class_results_with_corrections, 'output/PUP_count_correlation_classintake.csv', row.names = FALSE)
```


## Compound Modelling
```{r}
filtered_genus_long_compound = filtered_genus_long %>%
  left_join(compound_filtered, by = c("subject_id" = "UserName"))
```

Run LOOP
```{r}
# Create empty lists to store results and skipped combinations
compound_results_list = list()

compound_list = colnames(compound_filtered)[2:length(compound_filtered)]

# Run model for each polyphenol compound
for (compound_i in compound_list) {
  
  message("Running models for compound: ", compound_i)

  for (i in seq_len(nrow(gene_genus_combo))) {
    
    gene_i = gene_genus_combo$gene[i]
    pup_i = gene_genus_combo$PUPname[i]
    genus_i = gene_genus_combo$genus[i]
  
  message("Running model ", i, ": gene = ", gene_i, ", PUPname = ", pup_i, ", genus = ", genus_i)
  
    # Subset the data for each unique genus and gene
    subset = filtered_genus_long_compound %>%
    filter(gene == gene_i & PUPname == pup_i & genus == genus_i) 
    
    # Pull in compound data
    select_compound = subset[[compound_i]]
    
     # Perform Spearman correlation
    cor_test = cor.test(subset$read_count_sum, select_compound,
                        method = "spearman", exact = FALSE)
    
    output = tibble(
      gene = gene_i,
      PUPname = pup_i,
      genus = genus_i,
      compound = compound_i, 
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
    
    # Add formatted output to a list
    compound_results_list[[length(compound_results_list) + 1]] <- output
  }
}
```

Combine and format model output
```{r}
# Combine results into a tibble
compound_results = bind_rows(compound_results_list)

# Remove poor models before applying multiple comparison adjustment
compound_results_with_corrections = compound_results %>%
  group_by(gene) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(compound_results_with_corrections, 'output/PUP_count_correlation_compoundintake.csv', row.names = FALSE)
```