---
title: "PUP_Correlation_Abundance"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

# SUMMARY
Assessing the impact of polyphenol intake (total, class, compound) on microbial abundance using Spearman Correlation.

## INPUT
PUP_taxonomy_modelling_data.Rdata
  - meta: FL100 participant data including age, BMI, sex, total fiber intake, total HEI score, and total polyphenol intake
  - PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
  - tax: microbial PUP count data in long form
  - gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
  - class_filtered: class intakes that were in at least 30% of our cohort
  - compound_filtered: compound intakes that were in at least 30% of our cohort
  
genus_level_abundance.RData
  - genus_input: abundance of genera within FL100

## OUTPUT
  - **PUP_abundance_correlation_totalintake.csv**
  - **PUP_abundance_correlation_classintake.csv**
  - **PUP_abundance_correlation_compoundintake.csv**

## DATA PREP  
Load packages
```{r}
library(tidyverse)
```

Load Data
```{r}
load("RData/PUP_taxonomy_modelling_data.Rdata")

# From A0b_microbiome_prep
load("RData/genus_level_abundance.RData")
```

## Filter microbial genus data
First create genus list and search pattern
```{r}
genus_list = gene_genus_combo %>%
  distinct(genus) %>%
  mutate(search_term = str_remove(genus, "^g__"))

pattern = str_c(genus_list$search_term, collapse = "|")
```

Filter our microbial abundance data by the pattern
```{r}
filtered_genus_long = genus_input %>%
  filter(str_detect(Genus, pattern)) %>%
  filter(!str_detect(Genus, "unclassified")) %>%
  # Make it long format
  pivot_longer(-Genus, names_to = "subject_id") %>%
  mutate(subject_id = as.integer(subject_id)) %>%
  # Merge participant data
  filter(subject_id %in% meta$subject_id) %>%
  left_join(meta)

final_genus_list = genus_list %>% 
  filter(genus %in% filtered_genus_long$Genus)
```

## Total Modelling
```{r}
# Create empty lists to store results and skipped combinations
total_results_list = list()

for (i in seq_len(nrow(final_genus_list))) {
  
  genus_i = final_genus_list$genus[i]
  
  message("Running model ", i, ": genus = ", genus_i)
  
  subset = filtered_genus_long %>%
    filter(Genus == genus_i) 
  
    # Perform Spearman correlation
    cor_test = cor.test(subset$value, subset$ppintake_mg1000kcal,
                        method = "spearman", exact = FALSE)
    
    # Format output
    output = tibble(
      genus = genus_i,
      Variable = "ppintake_mg1000kcal",
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
  
  total_results_list[[length(total_results_list) + 1]] <- output
}
```

Combine and format model output 
```{r}
# Combine results into a tibble
total_results = bind_rows(total_results_list)

total_results_with_corrections = total_results %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(total_results_with_corrections, 'output/PUP_abundance_correlation_totalintake.csv', row.names = FALSE)
```


## Class Modelling
```{r}
filtered_genus_long_class = filtered_genus_long %>%
  left_join(class_filtered, by = c("subject_id" = "UserName"))
```

Run LOOP
```{r}
# Create empty lists to store results and skipped combinations
class_results_list = list()

class_list = colnames(class_filtered)[2:length(class_filtered)]

for (class_i in class_list) {
  message("Running models for class: ", class_i)

  for (i in seq_len(nrow(final_genus_list))) {
    
    genus_i = final_genus_list$genus[i]
    
    message("  Running model ", i, ": genus = ", genus_i)
    
    subset = filtered_genus_long_class %>%
      filter(Genus == genus_i)
    
    # Pull in class data
    select_class = subset[[class_i]]
    
     # Perform Spearman correlation
    cor_test = cor.test(subset$value, select_class,
                        method = "spearman", exact = FALSE)
    
    output = tibble(
      genus = genus_i,
      class = class_i, 
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
    
    class_results_list[[length(class_results_list) + 1]] <- output
  }
}
```

Combine and format modelling output
```{r}
# Combine results into a tibble
class_results = bind_rows(class_results_list)

class_results_with_corrections = class_results %>%
  group_by(genus) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(class_results_with_corrections, 'output/PUP_abundance_correlation_classintake.csv', row.names = FALSE)
```


## Compound Modelling
```{r}
filtered_genus_long_compound = filtered_genus_long %>%
  left_join(compound_filtered, by = c("subject_id" = "UserName"))
```

Run LOOP
```{r}
# Create empty lists to store results and skipped combinations
compound_results_list = list()

compound_list = colnames(compound_filtered)[2:length(compound_filtered)]

for (compound_i in compound_list) {
  message("Running models for compound: ", compound_i)

  for (i in seq_len(nrow(final_genus_list))) {
    
    genus_i = final_genus_list$genus[i]
    
    message("  Running model ", i, ": genus = ", genus_i)
    
    subset = filtered_genus_long_compound %>%
      filter(Genus == genus_i)
    
    # Pull in class data
    select_compound = subset[[compound_i]]
    
     # Perform Spearman correlation
    cor_test = cor.test(subset$value, select_compound,
                        method = "spearman", exact = FALSE)
    
    output = tibble(
      genus = genus_i,
      compound = compound_i, 
      estimate = cor_test$estimate,
      p_value = cor_test$p.value
    )
    
    compound_results_list[[length(compound_results_list) + 1]] <- output
  }
}
```

Combine and format model output
```{r}
# Combine results into a tibble
compound_results = bind_rows(compound_results_list)

compound_results_with_corrections = compound_results %>%
  group_by(genus) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) 

write.csv(compound_results_with_corrections, 'output/PUP_abundance_correlation_compoundintake.csv', row.names = FALSE)
```

