---
title: "E2_PUP_Modelling_Abundance"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

# SUMMARY
Assessing the impact of polyphenol intake (total, class, compound) on microbial abundance using  rank based estimation regression loops.

## INPUT
PUP_taxonomy_modelling_data.Rdata
  - meta: FL100 participant data including age, BMI, sex, total fiber intake, total HEI score, and total polyphenol intake
  - PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
  - tax: microbial PUP count data in long form
  - gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
  - class_filtered: class intakes that were in at least 30% of our cohort
  - compound_filtered: compound intakes that were in at least 30% of our cohort
  
genus_level_abundance.RData
  - genus_input: abundance of genera within FL100

## OUTPUT
  - PUP_abundance_modelling_totalintake.csv
  - PUP_abundance_modelling_classintake.csv
  - PUP_abundance_modelling_compoundintake.csv
  
Load packages
```{r}
library(tidyverse); library(Rfit)
```

Load Data
```{r}
load("RData/PUP_taxonomy_modelling_data.Rdata")

# From A0b_microbiome_prep
load("RData/genus_level_abundance.RData")
```

## Filter microbial genus data
First create genus list and search pattern
```{r}
genus_list = gene_genus_combo %>%
  distinct(genus) %>%
  mutate(search_term = str_remove(genus, "^g__"))

pattern = str_c(genus_list$search_term, collapse = "|")
```

Filter our microbial abundance data by the pattern
```{r}
filtered_genus_long = genus_input %>%
  filter(str_detect(Genus, pattern)) %>%
  filter(!str_detect(Genus, "unclassified")) %>%
  # Make it long format
  pivot_longer(-Genus, names_to = "subject_id") %>%
  mutate(subject_id = as.integer(subject_id)) %>%
  # Merge participant data
  filter(subject_id %in% meta$subject_id) %>%
  left_join(meta)

final_genus_list = genus_list %>% 
  filter(genus %in% filtered_genus_long$Genus)
```

## Total Modelling
```{r}
# Create empty lists to store results and skipped combinations
total_results_list = list()

for (i in seq_len(nrow(final_genus_list))) {
  
  genus_i = final_genus_list$genus[i]
  
  message("Running model ", i, ": genus = ", genus_i)
  
  subset = filtered_genus_long %>%
    filter(Genus == genus_i) 
  
  model = rfit(value ~ ppintake_mg1000kcal + Age + Sex + BMI + 
                 fiber_g1000kcal + hei_asa24_totalscore, 
               data = subset)
  
  output = as.data.frame(summary(model)$coefficients) %>%
    rownames_to_column(var = "Variable") %>%
    mutate(genus = genus_i) %>%
    relocate(genus, .before = Variable)
  
  total_results_list[[length(total_results_list) + 1]] <- output
}
```

Combine and format model output 
```{r}
# Combine results into a tibble
total_results = bind_rows(total_results_list)

total_results_with_corrections = total_results %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) 

write.csv(total_results_with_corrections, 'output/PUP_abundance_modelling_totalintake.csv', row.names = FALSE)
```


## Class Modelling
```{r}
filtered_genus_long_class = filtered_genus_long %>%
  left_join(class_filtered, by = c("subject_id" = "UserName"))
```

Run LOOP
```{r}
# Create empty lists to store results and skipped combinations
class_results_list = list()

class_list = colnames(class_filtered)[2:length(class_filtered)]

for (class_i in class_list) {
  message("Running models for class: ", class_i)

  for (i in seq_len(nrow(final_genus_list))) {
    
    genus_i = final_genus_list$genus[i]
    
    message("  Running model ", i, ": genus = ", genus_i)
    
    subset = filtered_genus_long_class %>%
      filter(Genus == genus_i)
    
    # Dynamically build formula with class_i as variable name
    formula_str = paste0(
      "value ~ `", class_i, "` + Age + Sex + BMI + fiber_g1000kcal + hei_asa24_totalscore"
    )
    
    model = rfit(
      formula = as.formula(formula_str),
      data = subset
    )
    
    output = as.data.frame(summary(model)$coefficients) %>%
      rownames_to_column(var = "Variable") %>%
      mutate(genus = genus_i,
             class = class_i) %>%
      relocate(genus, class, .before = Variable)
    
    class_results_list[[length(class_results_list) + 1]] <- output
  }
}
```

Combine and format modelling output
```{r}
# Combine results into a tibble
class_results = bind_rows(class_results_list)

class_results_with_corrections = class_results %>%
  group_by(genus) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) 

write.csv(class_results_with_corrections, 'output/PUP_abundance_modelling_classintake.csv', row.names = FALSE)
```


## Compound Modelling
```{r}
filtered_genus_long_compound = filtered_genus_long %>%
  left_join(compound_filtered, by = c("subject_id" = "UserName"))
```

Run LOOP
```{r}
# Create empty lists to store results and skipped combinations
compound_results_list = list()

compound_list = colnames(compound_filtered)[2:length(compound_filtered)]

for (compound_i in compound_list) {
  message("Running models for compound: ", compound_i)

  for (i in seq_len(nrow(final_genus_list))) {
    
    genus_i = final_genus_list$genus[i]
    
    message("  Running model ", i, ": genus = ", genus_i)
    
    subset = filtered_genus_long_compound %>%
      filter(Genus == genus_i)
    
    # Dynamically build formula with compound_i as variable name
    formula_str = paste0(
      "value ~ `", compound_i, "` + Age + Sex + BMI + fiber_g1000kcal + hei_asa24_totalscore"
    )
    
    model = rfit(
      formula = as.formula(formula_str),
      data = subset
    )
    
    output = as.data.frame(summary(model)$coefficients) %>%
      rownames_to_column(var = "Variable") %>%
      mutate(genus = genus_i,
             compound = compound_i) %>%
      relocate(genus, compound, .before = Variable)
    
    compound_results_list[[length(compound_results_list) + 1]] <- output
  }
}
```

Combine and format model output
```{r}
# Combine results into a tibble
compound_results = bind_rows(compound_results_list)

compound_results_with_corrections = compound_results %>%
  group_by(genus) %>%
  mutate(p_adj = p.adjust(p.value, method = "BH")) 

write.csv(compound_results_with_corrections, 'output/PUP_abundance_modelling_compoundintake.csv', row.names = FALSE)
```

