---
title: "PUP_Modelling_Summary"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

## SUMMARY
This script merges total, class, and compound-level model output files into one singular file. This is done for both models on microbial abundance and PUP-gene counts.

## INPUT
PUP_taxonomy_modelling_data.Rdata
  - meta: FL100 participant data including age, BMI, sex, total fiber intake, total HEI score, and total polyphenol intake
  - PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
  - tax: microbial PUP count data in long form
  - gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
  - class_filtered: class intakes that were in at least 30% of our cohort
  - compound_filtered: compound intakes that were in at least 30% of our cohort
  
Count Modelling Outputs
  - PUP_count_modelling_totalintake.csv
  - PUP_count_modelling_classintake.csv
  - PUP_count_modelling_compoundintake.csv

Abundance Modelling Outputs
  - PUP_abundance_modelling_totalintake.csv
  - PUP_abundance_modelling_classintake.csv
  - PUP_abundance_modelling_compoundintake.csv

## OUTPUT  
  - *PUP_Modelling_Summary.xlsx* - File with total, class, and compound-level model outputs merged into long-format

## DATA PREP
Load packages
```{r}
library(tidyverse); library(writexl)
```

Load Data
```{r}
load("RData/PUP_taxonomy_modelling_data.Rdata")

# Load model outputs
total_count = read.csv('output/PUP_count_modelling_totalintake.csv') %>%
  mutate(resolution = "total")
class_count = read.csv('output/PUP_count_modelling_classintake.csv') %>%
  mutate(resolution = "class")
compound_count = read.csv('output/PUP_count_modelling_compoundintake.csv') %>%
  mutate(resolution = "compound")

total_abund = read.csv('output/PUP_abundance_modelling_totalintake.csv') %>%
  mutate(resolution = "total")
class_abund = read.csv('output/PUP_abundance_modelling_classintake.csv') %>%
  mutate(resolution = "class")
compound_abund = read.csv('output/PUP_abundance_modelling_compoundintake.csv') %>%
  mutate(resolution = "compound")
```

```{r}
substrate_combo = read.csv('dbPUP/polyphenol_substrates_long_FooDB_taxonomy.csv') %>%
  rename(Variable = compound_public_id)

# Load substrate information
substrate_list = substrate_combo %>%
  select(c(Variable, compound_name)) %>%
  filter(!is.na(Variable)) %>% # Remove missing values
  # So this matches our model output
  distinct(Variable, .keep_all = TRUE) %>% # Remove duplicates
  mutate(known_substrate = "yes")

polyphenol_list = read.csv('data/FooDB_phenols_taxonomy.csv') %>%
  rename(Variable = compound_public_id) %>%
  select(c(Variable, compound_name)) %>% 
  filter(!Variable %in% substrate_list$Variable) %>%
  mutate(known_substrate = "no")

all_polyphenols = full_join(substrate_list, polyphenol_list)
```

## Combine Model Outputs
```{r}
count_merge = full_join(total_count, class_count) %>%
  full_join(compound_count)

abund_merge = full_join(total_abund, class_abund) %>%
  full_join(compound_abund) 
```

## Create list of polyphenol resolution variables
```{r}
total = "ppintake_mg1000kcal"
class = colnames(class_filtered)[-1]
compound = colnames(compound_filtered)[-1]

# merge into one list
polyphenol_terms = c(total, class, compound)
```

## Signficant polyphenol findings
```{r}
count_poly_sig = count_merge %>%
  # Filter significant polyphenol intake findings
  filter(p_adj <= 0.05 & Variable %in% polyphenol_terms) %>%
  select(-c(class, compound)) %>%
  left_join(all_polyphenols) %>%
  mutate(resolution = factor(resolution, levels = c("total", "class", "compound")),
         PUP_Uniprot = paste(PUPname, " (", gene, ")", sep = ""),
         PUP_Uniprot_v2 = paste(PUPname, gene, sep = ", "),
         Genus = str_remove_all(genus, 'g__'),
         Estimate = round(Estimate, 3),
         p_adj = ifelse(p_adj < 0.001, "<0.001", round(p_adj, 3)),
         # Indicate gene and substrate combinations that were in dbPUP
         Substrate = if_else(map2_lgl(gene, Variable, ~ 
                                      any(.x == substrate_combo$gene & .y == substrate_combo$Variable)), "*", ""),
         Polyphenol = if_else(str_starts(Variable, "FDB"),
                              if_else(Substrate == "*", 
                                      paste0(compound_name, " (", Variable, ")*"), 
                                      paste0(compound_name, " (", Variable, ")")), Variable)) %>%
  select(c(resolution, Genus, PUP_Uniprot, PUP_Uniprot_v2, Polyphenol, Estimate, p_adj)) %>%
  arrange(resolution, Genus, PUP_Uniprot)

abund_poly_sig = abund_merge %>%
  # Filter significant polyphenol intake findings
  filter(p_adj <= 0.05 & Variable %in% polyphenol_terms) %>%
  select(-c(class, compound)) %>%
  left_join(all_polyphenols) %>%
  mutate(resolution = factor(resolution, levels = c("total", "class", "compound")),
         Genus = str_remove_all(genus, 'g__'),
         Estimate = round(Estimate, 3),
         p_adj = ifelse(p_adj < 0.001, "<0.001", round(p_adj, 3)),
         Polyphenol = if_else(str_starts(Variable, "FDB"),
                              if_else(known_substrate == "yes",
        paste0(compound_name, " (", Variable, ")*"), 
        paste0(compound_name, " (", Variable, ")")), Variable)) %>%
  select(c(resolution, Genus, Polyphenol, Estimate, p_adj)) %>%
  arrange(resolution, Genus)
```

Save output into multi-tab Excel file
```{r}
write_xlsx(list(count_models_all = count_merge,
                abund_models_all = abund_merge,
                count_models_polyphenol = count_poly_sig,
                abund_models_polyphenols = abund_poly_sig),
           path = "output/PUP_Modelling_Summary.xlsx")
```