---
title: "PUP_Modelling_Summary_Covariates"
author: "Stephanie Wilson"
date: "April 2026"
output: html_document
---

## SUMMARY
This script merges total, class, and compound-level model output files into one singular file. This is done for both models on microbial abundance and PUP-gene counts. We pull out the **covariate** impacts here for the creation of Supplemental Tables 2 and 4. 

## INPUT
PUP_taxonomy_modelling_data.Rdata
  - meta: FL100 participant data including age, BMI, sex, total fiber intake, total HEI score, and total polyphenol intake
  - PUPnames: minimal PUP name dataframe including gene, PUPname, PUPclass
  - tax: microbial PUP count data in long form
  - gene_genus_combo: unique combinations of genus (with at least 10% prevalence) and PUP genes
  - class_filtered: class intakes that were in at least 30% of our cohort
  - compound_filtered: compound intakes that were in at least 30% of our cohort
  
Count Modelling Outputs
  - PUP_count_modelling_totalintake.csv
  - PUP_count_modelling_classintake.csv
  - PUP_count_modelling_compoundintake.csv

Abundance Modelling Outputs
  - PUP_abundance_modelling_totalintake.csv
  - PUP_abundance_modelling_classintake.csv
  - PUP_abundance_modelling_compoundintake.csv

## OUTPUT  
  - *PUP_Modelling_Summary_Covariates.xlsx* - File with total, class, and compound-level model outputs merged into long-format, 

## DATA PREP
Load packages
```{r}
library(tidyverse); library(writexl)
```

Load Data
```{r}
load("RData/PUP_taxonomy_modelling_data.Rdata")

# Load model outputs
total_count = read.csv('output/PUP_count_modelling_totalintake.csv') %>%
  mutate(resolution = "total")
class_count = read.csv('output/PUP_count_modelling_classintake.csv') %>%
  mutate(resolution = "class")
compound_count = read.csv('output/PUP_count_modelling_compoundintake.csv') %>%
  mutate(resolution = "compound")

total_abund = read.csv('output/PUP_abundance_modelling_totalintake.csv') %>%
  mutate(resolution = "total")
class_abund = read.csv('output/PUP_abundance_modelling_classintake.csv') %>%
  mutate(resolution = "class")
compound_abund = read.csv('output/PUP_abundance_modelling_compoundintake.csv') %>%
  mutate(resolution = "compound")

# Load taxonomy data
substrate_combo = read.csv('dbPUP/polyphenol_substrates_long_FooDB_taxonomy.csv') %>%
  rename(Variable = compound_public_id)
```

```{r}
# Load substrate information
substrate_list = substrate_combo %>%
  select(c(Variable, compound_name)) %>%
  filter(!is.na(Variable)) %>% # Remove missing values
  # So this matches our model output
  distinct(Variable, .keep_all = TRUE) %>% # Remove duplicates
  mutate(known_substrate = "yes")

polyphenol_list = read.csv('data/FooDB_phenols_taxonomy.csv') %>%
  rename(Variable = compound_public_id) %>%
  select(c(Variable, compound_name)) %>% 
  filter(!Variable %in% substrate_list$Variable) %>%
  mutate(known_substrate = "no")

all_polyphenols = full_join(substrate_list, polyphenol_list)
```

## Combine Model Outputs
```{r}
count_merge = full_join(total_count, class_count) %>%
  full_join(compound_count) %>%
  left_join(all_polyphenols) %>%
  # Format variables to be publication ready
  mutate(resolution = factor(resolution, levels = c("total", "class", "compound")),
         PUP_Uniprot = paste(PUPname, " (", gene, ")", sep = ""),
         PUP_Uniprot_v2 = paste(PUPname, gene, sep = ", "),
         Genus = str_remove_all(genus, 'g__'),
         Estimate = round(Estimate, 3),
         p_adj = ifelse(p_adj < 0.001, "<0.001", round(p_adj, 3)),
         Variable = case_when(
           Variable == "BMI" ~ "BMI (kg/m2)",
           Variable == "SexMale" ~ "Sex, Male",
           Variable == "fiber_g1000kcal" ~ "Fiber (g/1000 kcal)",
           Variable == "hei_asa24_totalscore" ~ "Total HEI-2015 Score",
            .default = Variable),
          resolution_info = coalesce(class, compound)) 

abund_merge = full_join(total_abund, class_abund) %>%
  full_join(compound_abund) %>%
  # Format variables to be publication ready
  mutate(resolution = factor(resolution, levels = c("total", "class", "compound")),
         Genus = str_remove_all(genus, 'g__'),
         Estimate = round(Estimate, 3),
         p_adj = ifelse(p_adj < 0.001, "<0.001", round(p_adj, 3)),
         Variable = case_when(
           Variable == "BMI" ~ "BMI (kg/m2)",
           Variable == "SexMale" ~ "Sex, Male",
           Variable == "fiber_g1000kcal" ~ "Fiber (g/1000 kcal)",
           Variable == "hei_asa24_totalscore" ~ "Total HEI-2015 Score",
           .default = Variable),
         resolution_info = coalesce(class, compound)) %>%
  select(c(resolution, resolution_info, Genus, Variable, Estimate, p_adj))
```


## Signficant polyphenol findings
```{r}
count_cov_sig = count_merge %>%
  # Filter significant covariate findings
  filter(p_adj <= 0.05 & Variable %in% c("Age", "BMI (kg/m2)", "Sex, Male", 
                                         "Fiber (g/1000 kcal)", "Total HEI-2015 Score")) %>%
  group_by(resolution, Genus, PUP_Uniprot, Variable) %>%
  # Summarize covariate and p-values across models
  summarise(
    n_unique = n_distinct(resolution_info),
    estimate_min = min(Estimate, na.rm = TRUE),
    estimate_max = max(Estimate, na.rm = TRUE),
    p_adj_min = min(p_adj, na.rm = TRUE),
    p_adj_max = max(p_adj, na.rm = TRUE),
    .groups = "drop")%>%
  relocate(n_unique, .after = resolution)

abund_cov_sig = abund_merge %>%
  # Filter significant covariate findings
    filter(p_adj <= 0.05 & Variable %in% c("Age", "BMI (kg/m2)", "Sex, Male", 
                                         "Fiber (g/1000 kcal)", "Total HEI-2015 Score")) %>%
  # Summarize covariate and p-values across models
  group_by(resolution, Genus, Variable) %>%
  summarise(
    n_unique = n_distinct(resolution_info),
    estimate_min = min(Estimate, na.rm = TRUE),
    estimate_max = max(Estimate, na.rm = TRUE),
    p_adj_min = min(p_adj, na.rm = TRUE),
    p_adj_max = max(p_adj, na.rm = TRUE),
    .groups = "drop") %>%
  relocate(n_unique, .after = resolution)
```

Save output into multi-tab Excel file
```{r}
write_xlsx(list(count_models_covariates = count_cov_sig,
                abund_models_covariates = abund_cov_sig),
           path = "output/PUP_Modelling_Summary_Covariates.xlsx")
```