---
title: "Compare_Correlation_Rfits"
author: "Stephanie Wilson"
date: "June 2025"
output: html_document
---

## SUMMARY
Compares Correlation (covariate unadjusted) vs Rank-based estimation regression (covariate adjusted) outputs.

## INPUT
  - PUP_Modelling_Summary.xlsx - Combined outputs from Rfit models
  - PUP_Correlation_Summary.xlsx - combined outputs from correlations
  
## OUTPUT
  - No files
  
## PREP  
Load packages
```{r}
library(tidyverse); library(readxl)
```

Load Data
```{r}
# PUP gene class info
class_info = read.csv("dbPUP/dbpup_overview_clean.csv") %>%
  select(c(gene, class, Family.Number)) 

count_corr = read_xlsx("output/PUP_Correlation_Summary.xlsx", sheet = 3)
abundance_corr= read_xlsx("output/PUP_Correlation_Summary.xlsx", sheet = 4)

count_modelling = read_xlsx("output/PUP_Modelling_Summary.xlsx", sheet = 3) 
abundance_modelling = read_xlsx("output/PUP_Modelling_Summary.xlsx", sheet = 4)
```

Prep data file by combining PUP class information
```{r}
output_class = count_modelling %>%
  mutate(gene = str_extract(PUP_Uniprot, "(?<=\\()[^)]*(?=\\))")) %>%
  left_join(class_info) %>%
  # Prepare more informative name
  mutate(PUP_Uniprot = case_when(
          PUP_Uniprot == "RhaB2 (C4PG45)" ~ "Alpha-L-rhamnosidase B2 (C4PG45)",
          .default = PUP_Uniprot))
```


## Are there relationships shared between our correlation and rank-based regression for microbial PUP counts?
```{r}
shared_count_keys = intersect(
  count_corr %>% select(resolution, Genus, PUP_Uniprot, Polyphenol) %>% distinct(),
  count_modelling %>% select(resolution, Genus, PUP_Uniprot, Polyphenol) %>% distinct())

shared_count_keys
```
There are no shared entries for microbial PUP gene counts. 

## Are there relationships shared between our correlation and rank-based regression for microbial abundance counts?
```{r}
shared_abundance_keys = intersect(
  abundance_corr %>% select(resolution, Genus, Polyphenol) %>% distinct(),
  abundance_modelling %>% select(resolution, Genus,  Polyphenol) %>% distinct())

shared_abundance_keys
```
Only one relationship was detected with and without covariate in modelling. 

## What classes are most represented?
```{r}
output_class %>%
  distinct(PUP_Uniprot, .keep_all = TRUE) %>%
  group_by(class) %>%
  summarise(count = n())

output_class %>%
  group_by(class) %>%
  summarise(count = n())
```

## What Genera are most represented?
```{r}
output_class %>%
  group_by(Genus) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:5)
```

## Of the compounds that show up the most, what are the major linked PUP genes?
```{r}
most_genus = output_class %>%
  filter(grepl("Apigenin|Daidzein|Diosmetin 7-rutinoside", Polyphenol)) 

output_class %>%
  filter(grepl("Alpha-L-rhamnosidase|Beta-glucosidase", PUP_Uniprot)) %>%
  group_by(PUP_Uniprot) %>%
  summarize(count = n())
```

## Since hydrolysis PUP genes were most common, which microbial genera were these attached to?
```{r}
output_class %>%
  filter(class == "hydrolysis") %>%
  group_by(Genus) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```

