---
title: "Modelling_Overview_PLOTS"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

## Summary
Create publication-ready heatmap visuals for the modelling polyphenol outputs. These are main-text manuscript figures.

## INPUT
  - PUP_Modelling_Summary.xlsx
  - dbpup_overview_clean.csv
  
## OUTPUT
Heatmap image files for PUP count and microbial abundance data
- Abundance_Modelling_plot, PDF + PNG
- count_model_plot_data.csv
- Count_Modelling_plot.png

## PREP  
Load packages
```{r warning = FALSE, message = FALSE}
library(readxl); library(tidyverse); library(scales); library(ggtext)
```

Load data
```{r}
# PUP gene class info
class_info = read.csv("dbPUP/dbpup_overview_clean.csv") %>%
  select(c(gene, class, Family.Number)) %>%
  mutate(class = case_when(
    class == "hydrolysis" ~ "HR",
    class == "functional_group_transfer" ~ "FR",
    class == "oxidation_reduction" ~ "OR",
    class == "unclassified" ~ "UC",
    .default = class
  ))

count_modelling = read_xlsx("output/PUP_Modelling_Summary.xlsx", sheet = 3)
abundance_modelling = read_xlsx("output/PUP_Modelling_Summary.xlsx", sheet = 4)
```

## Abundance Model Plot

Format labels to be more informative
```{r}
abundance_model_data = abundance_modelling %>%
  mutate(p_adj = as.numeric(str_replace(p_adj, "<0.001", "0.0009")),
         Substrate = if_else(str_detect(Polyphenol, "\\*"), "*", ""),
         Polyphenol = str_remove(Polyphenol, " \\(.*\\)"),
         # Polyphenol = str_remove(Polyphenol, "\\*"),
         Polyphenol = case_when(
           Polyphenol == "ppintake_mg1000kcal" ~ "Total Polyphenol",
           Polyphenol == "Flavanols_aka_flavan3ols" ~ "Flavanols (Flavan-3-ols)",
           .default = Polyphenol),
         resolution = factor(resolution, levels = c("total", "class", "compound")),
         Genus = paste0("italic('", Genus, "')"),
         Genus = fct_rev(Genus),
         squished = case_when(
           Estimate > 3 ~ "Above",
           Estimate < -3~ "Below",
           TRUE ~ NA_character_))
```

```{r}
abundance_model_plot = ggplot(abundance_model_data, aes(x = Polyphenol, y = Genus, 
                                                        fill = Estimate)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#500000", mid = "#D1D1D1", high = "#2D5D7B", 
                       midpoint = 0, 
                       name = "Estimate",
                       limits = c(-3,3),
                       oob = squish) +
  geom_point(data = filter(abundance_model_data, !is.na(squished)),
             aes(x = Polyphenol, y = Genus, shape = squished),
             color = "white", size = 2) +
  # Specify shapes for outlying values
  scale_shape_manual(values = c("Below" = 25, "Above" = 24),
                     name = "Outside\nDisplay\nRange") + # triangle down/up
  labs(x = "", y = "") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    panel.spacing = unit(0.0, "lines"),
    legend.position = "right") +
  facet_grid(. ~ resolution, scales = "free", space = "free") +
  guides(shape = guide_legend(override.aes = list(color = "black")),
         fill = guide_colorbar(order = 1),    # color gradient first
         shape = guide_legend(order = 2)) +      # outlier shapes second)
  scale_y_discrete(labels = scales::label_parse()) 

abundance_model_plot

ggsave("images/Abundance_Modelling_plot.pdf", width = 6, height = 4)
ggsave("images/Abundance_Modelling_plot.png", width = 6, height = 4, units = "in", dpi = 300)
```


## DIET VS PUP COUNT PLOT
Pull count modelling data and clean up labels for plotting
```{r}
count_model_data = count_modelling %>%
  # Add PUP class information
  mutate(gene = str_extract(PUP_Uniprot, "(?<=\\()[^)]*(?=\\))")) %>%
  left_join(class_info) %>%
  mutate(p_adj = as.numeric(str_replace(p_adj, "<0.001", "0.0009")),
         Substrate = if_else(str_detect(Polyphenol, "\\*"), "*", ""),
         Polyphenol = str_remove(Polyphenol, " \\(.*\\)"),
         Polyphenol = str_remove(Polyphenol, "\\*"),
         Polyphenol = case_when(
           Polyphenol == "ppintake_mg1000kcal" ~ "Total Polyphenol",
           Polyphenol == "Chalcones_Dihydrochalcones" ~ "Chalcones",
           .default = Polyphenol),
         resolution = factor(resolution, levels = c("total", "class", "compound")),
         # Lengthen Abbreviated Name
         PUP_Uniprot_v2 = case_when(
            PUP_Uniprot_v2 == "RhaB2 (C4PG45)" ~ "Alpha-L-rhamnosidase B2 (C4PG45)",
            .default = PUP_Uniprot_v2),
        # Replace alpha/beta with Greek letters
        PUP_Uniprot_v2 = str_replace_all(PUP_Uniprot_v2, 
                                      c("Alpha" = "α", "Beta" = "β", "beta" = "β")),
          # Add Reaction class abbreviation to the PUP name, Uniprot string
        PUP_Uniprot_v2 = paste(PUP_Uniprot_v2, class, sep = ", "),
        PUP_Uniprot_v2 = fct_rev(PUP_Uniprot_v2),
        Genus = paste0("italic('", Genus, "')"), # italicize genera
        squished = case_when(
           Estimate > 200 ~ "Above",
           Estimate < -200~ "Below",
           TRUE ~ NA_character_))

write.csv(count_model_data, "output/count_model_plot_data.csv", row.names = FALSE)
```

Create plot with x-axis as polyphenol intake across resolutions and y-axis as PUPs grouped by microbial genera. 
```{r}
count_model_plot = ggplot(count_model_data, aes(x = Polyphenol, y = PUP_Uniprot_v2, fill = Estimate)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#500000", mid = "#D1D1D1", high = "#2D5D7B", 
                       midpoint = 0, 
                       name = "Estimate",
                       limits = c(-200,200),
                       oob = squish) +
   geom_point(data = filter(count_model_data, !is.na(squished)),
             aes(x = Polyphenol, y = PUP_Uniprot_v2, shape = squished),
             color = "white", size = 2) +
  theme_bw() +
  facet_grid(Genus ~ resolution, 
             labeller = labeller(Genus = label_parsed), #italicize genus
             switch = "y", scales = "free", space = "free") +
  scale_y_discrete(position = "right") +  # Move y-axis ticks/labels to right
  scale_shape_manual(values = c("Below" = 25, "Above" = 24),
                     name = "Outside Display Range") +   # Specify shapes for outlying values
  labs(x = "", y = "") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y.right = element_text(size = 8),     # y-axis tick labels
    axis.text.y.left = element_blank(),             # hide default left ticks
    axis.title.y.left = element_blank(),            # hide default left label
    panel.spacing = unit(0.0, "lines"),             # reduce facet spacing
    strip.text.y.left = element_text(angle = 0, hjust =1),  # left-side facet horizontal
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.position = c(1, -0.20),  # Moves legend under facet, near x-axis labels
    legend.justification = c(0, 0),  # Aligns legend to bottom-left
    legend.direction = "horizontal") +  # Makes legend horizontal
  geom_text(data = filter(count_model_data, Substrate == "*"),
            aes(label = "*"), color = "black", size = 5) +
  guides(fill = guide_colorbar(order = 1, title.position = "top", title.hjust = 0.5),
         shape = guide_legend(order = 2, override.aes = list(color = "black")),
         color = guide_legend(order =3)
         ) # Ensure outlier points are black in the legend

# PDF does not export greek symbols
ggsave("images/Count_Modelling_plot.png", width = 14, height = 8.5, units = "in", dpi = 300)
```
