---
title: "Participant Characteristics Table"
author: "Stephanie Wilson"
date: "August 2024"
output: html_document
---

## Summary
This code prepares the participant characteristics table (Table 1) for the polyphenol/microbiome manuscript. Scripts calculate the overall mean and standard error for continuous variables and the mean and standard error by polyphenol quartile. Frequency of males and females in the overall cohort and by quartile group are also compiled. 

## Input
  - **metadata_313.Rdata**: filtered for those with metagenomic and diet data (n=313), output from A0a_dietary_prep.R
  
## Output
  - **FL100_MB_participant_table.csv** -content data for participant table, needs additional visual formatting
  
## Load packages and data
Load Packages
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Load Metadata and Dietary Data
```{r message = FALSE}
load("RData/metadata_313.Rdata")

# Specify variables of interest for the table
continuous = c("Age", "BMI", "fiber_g1000kcal", "hei_asa24_totalscore", "ppintake_mg1000kcal", "waistavg_cm") # ,

#unicode 
plus_minus = "\u00B1"
```

## Overall summary

Sex count, Coffee Count
```{r}
overall_sex = data.frame(table(cleaned_data$Sex)) %>%
  pivot_wider(names_from = Var1, values_from = Freq) %>%
  mutate(variable = "Female/Male") %>%
  mutate(Overall = paste(Female, Male, sep = "/")) %>%
  select(-c(Female, Male)) 

overall_coffee = data.frame(table(cleaned_data$coffee_consumer)) %>%
  pivot_wider(names_from = Var1, values_from = Freq) %>%
  mutate(variable = "Coffee Consumer")  %>%
  mutate(Overall = paste(yes, no, sep = "/")) %>%
  select(-c(no, yes)) 
```

summary dataframe for 313 participants
```{r}
overall_summary = cleaned_data %>%
  summarise(across(all_of(continuous), 
                   list(mean = ~ mean(. , na.rm = TRUE), 
                        se = ~ sd(. , na.rm = TRUE) / sqrt(n())))) %>%
  pivot_longer(cols = everything(), 
               names_to = c("variable", ".value"), 
               names_pattern = "(.*)_(.*)") %>%
  mutate(Overall = paste(round(mean, digits =1), 
                         plus_minus, 
                         round(se, digits =1))) %>%
  select(-c(mean, se))
```

## BY Polyphenol Quartile

Counts by Quartile
```{r}
subset = cleaned_data %>%
     filter(pp_quartile_label %in% c("Low", "High"))

quartile_sex = data.frame(table(subset$Sex, subset$pp_quartile_label)) %>%
  pivot_wider(names_from = Var1, values_from = Freq) %>%
  mutate(FMratio = paste(Female, Male, sep = "/")) %>%
  select(-c(Female, Male)) %>%
  t() %>%
  as_tibble() %>%
  mutate(variable = "Female/Male") %>%
  rename(High = V1, Low = V2) %>%
  slice(-1)

quartile_coffee = data.frame(table(subset$coffee_consumer, subset$pp_quartile_label)) %>%
  pivot_wider(names_from = Var1, values_from = Freq) %>%
  mutate(yes_no = paste(yes, no, sep = "/")) %>%
  select(-c(yes, no)) %>%
  t() %>%
  as_tibble() %>%
  mutate(variable = "Coffee Consumer") %>%
  rename(High = V1, Low = V2) %>%
  slice(-1)
```

Generate quartile summary
```{r}
quartile_summary = cleaned_data %>%
  filter(pp_quartile_label %in% c("Low", "High")) %>%
  group_by(pp_quartile_label) %>%
  summarise(across(all_of(continuous), 
                   list(mean = ~ mean(. , na.rm = TRUE), 
                        se = ~ sd(. , na.rm = TRUE) / sqrt(n())))) %>%
    pivot_longer(cols = -pp_quartile_label, 
               names_to = c("variable", ".value"), 
               names_pattern = "(.+)_(.+)") %>%
  mutate(overall = paste(round(mean, digits =1), 
                         plus_minus, 
                         round(se, digits =1))) %>%
  select(-c(mean, se)) %>%
  pivot_wider(names_from = pp_quartile_label, values_from = overall)
```

## Statistical Testing BY Polyphenol Quartile
```{r}
# Initialize an empty dataframe to store the t-test results
t_test_results_df = data.frame(
  variable = character(),
  statistic = numeric(),
  p = numeric(),
  Mean_Low = numeric(),
  Mean_High = numeric(),
  stringsAsFactors = FALSE)

# Loop through each continuous variable and perform a t-test
for (var in continuous) {
  
  # Perform the t-test
  t_test_result = t.test(subset[[var]] ~ subset$pp_quartile_label)
  
  # Extract the means for each group
  means = subset %>%
    group_by(pp_quartile_label) %>%
    dplyr::summarize(mean_value = mean(get(var), na.rm = TRUE, .groups = 'drop'))
  
  mean_low = means %>% filter(pp_quartile_label == "Low") %>% pull(mean_value)
  mean_high = means %>% filter(pp_quartile_label == "High") %>% pull(mean_value)
  
  # Add the results to the dataframe
  t_test_results_df = rbind(t_test_results_df, data.frame(
    variable = var,
    statistic = t_test_result$statistic,
    p = t_test_result$p.value,
    Mean_Low = mean_low,
    Mean_High = mean_high,
    stringsAsFactors = FALSE))
}
```

Statistics for count data
```{r}
contingency_table_sex = table(subset$pp_quartile_label, subset$Sex)
contingency_table_coffee = table(subset$pp_quartile_label, subset$coffee_consumer)

# Perform the chi-square test, sex
chi_square_test_sex = chisq.test(contingency_table_sex)
chi_sum_sex = data.frame(variable = 'Female/Male', 
                        statistic = chi_square_test_sex$statistic,
                        p = chi_square_test_sex$p.value, row.names = NULL)

# Perform the chi-square test, coffee
chi_square_test_coffee = chisq.test(contingency_table_coffee)
chi_sum_coffee = data.frame(variable = 'Coffee Consumer', 
                        statistic = chi_square_test_coffee$statistic,
                        p = chi_square_test_coffee$p.value, row.names = NULL)

chi_sum_merge = full_join(chi_sum_sex, chi_sum_coffee)
```


```{r}
statistic_merge = full_join(chi_sum_merge, t_test_results_df) %>%
  select(-c(Mean_Low, Mean_High)) %>%
  mutate(statistic = round(statistic, digits =2))
```


## Merge Dataframes
```{r}
merge_sex = left_join(overall_sex, quartile_sex)
merge_coffee = left_join(overall_coffee, quartile_coffee)
merge_summary = left_join(overall_summary, quartile_summary) %>%
  select(c(variable, Overall, Low, High))
```

Polish Column Names
```{r}
new_overall = paste("Overall (n = ", nrow(cleaned_data), ")", sep = "")
new_high = paste("High (n = ", nrow(subset %>% filter(pp_quartile_label =="High")), ")", sep = "")
new_low = paste("Low (n = ", nrow(subset %>% filter(pp_quartile_label =="Low")), ")", sep = "")
```

```{r}
merge_final = full_join(merge_sex, merge_coffee) %>%
  full_join(merge_summary) %>%
  rename(!!new_overall := Overall,
         !!new_high := High,
         !!new_low := Low) %>%
  left_join(statistic_merge)

# Export
write.csv(merge_final, 'output/FL100_MB_participant_table.csv', row.names = FALSE)
```

