---
title: "LPS Taxa Modelling"
author: "Stephanie Wilson"
date: "January 2024"
output: html_document
---

## Summary

Examine the relationship between polyphenol intake (prenol lipids, phenylpropanoic acids) and the abundance of pro-inflammatory LPS (Gammaproteobacteria, Enterobacteriaceae) and major LPS-producers in the healthy human gut (Bacteroidales).

## Input

-   **metadata_313.csv** - meta data for subjects with diet and metagenome data
-   **LPS_related_bacteria_metagenomic.csv** - Abundances of select bacterial taxa relevant for LPS Binding Protein Analysis

## Output

-   Effect plots for polyphenol intake, including publication ready summary plot: LPS_producers_model_plot.png
-   Results in line model output.

## Load packages and data

Packages

```{r message=FALSE, warning=FALSE}
library(tidyverse); library(effects); library(car);  library(bestNormalize)
library(gamair); library(mgcv) # For GAM
library(Rfit) #For Rank-Based Estimation Regrsesion
library(ggeffects); library(ggpubr)

source("functions/annotate_lm_plot.R")
```

Metadata and Dietary Data

```{r message = FALSE}
load("RData/metadata_313.Rdata")

LPS_bacteria = read.csv('data/LPS_related_bacteria_metagenomic.csv')

cleaned_data = cleaned_data %>%
  mutate(entero_binomial = ifelse(f__Enterobacteriaceae > 0, 1, 0),
         Prenol_Phenylprop = Prenol.lipids + Phenylpropanoic.acids) %>%
  left_join(LPS_bacteria)
```

## 1) Exploratory

How many zeros are we looking at?

```{r}
message("Participants without Gammaprotebacteria: ", 
        sum(cleaned_data$c__Gammaproteobacteria== 0),
        "\nParticipants without Enterobacteriaceae: ", 
        sum(cleaned_data$f__Enterobacteriaceae== 0))
```

## 2) Modelling

### 2a) Rank-based Estimation Regression - Gammaproteobacteria

```{r}
gamma_transform = bestNormalize(cleaned_data$c__Gammaproteobacteria)
cleaned_data$gamma_log = gamma_transform$other_transforms$log_x$x.t

model_gamma_lm = lm(gamma_log ~ Prenol_Phenylprop + Sex + Age + BMI
                    + fiber_g1000kcal + hei_asa24_totalscore, data = cleaned_data %>% filter(!UserName == "7098"))

summary(model_gamma_lm)
resid_panel(model_gamma_lm)
```

### 2b) Logistic Regression - Enterobacteriaceae

Consumption of polyphenols considered phenylpropanoic acids and prenol lipids did not impact the probability of having Enterobacteriaceae (p = 0.41).

```{r}
mod_glm = glm(entero_binomial ~ Prenol_Phenylprop + Age + Sex + BMI +
                fiber_g1000kcal + hei_asa24_totalscore, family = "binomial", 
              data = cleaned_data)
summary(mod_glm)
```

### 2c) Bacteroidales

All participants have Bacteroidales, no zeros.

```{r}
bact_transform = bestNormalize(cleaned_data$Bacteroidales)
cleaned_data$Bacteroidales_yeo = bact_transform$x.t

mod_bacteroidales_lm  = lm(Bacteroidales_yeo ~ Prenol_Phenylprop + Sex + Age + BMI
                    + fiber_g1000kcal + hei_asa24_totalscore, data = cleaned_data)
summary(mod_bacteroidales_lm)

ggResidpanel::resid_panel(mod_bacteroidales_lm)
```

## Publication Ready Plots

### 2a 
```{r}
gamma_plot = annotate_lm_plot(model_gamma_lm, "Prenol_Phenylprop", 1.1) + 
  labs(
    x = "Intake of Prenol Lipids,\nPhenylpropanoic Acids\n(mg/1000 kcal)",
    y = expression(atop("Gammaproteobacteria", "Abundance (log"[10]*"-transformed)")))
```

### 2b
```{r}
glm_eff = ggpredict(mod_glm, terms = "Prenol_Phenylprop")

# Plot the effect
entero_plot = plot(glm_eff) +
  geom_jitter(data = mod_glm$data, aes(x = Prenol_Phenylprop, y = entero_binomial),
              width = 0, height = 0.02, alpha = 0.3) +
  labs(title = "",
       x = "Intake of Prenol Lipids,\nPhenylpropanoic Acids\n(mg/1000 kcal)",
       y = expression(Pr(italic(Enterobacteriaceae)))) +
  annotate("text", x = max(mod_glm$data$Prenol_Phenylprop) *0.9, 
                   y = 0.9,
             label = bquote(atop(beta == .(round(mod_glm$coefficients[2], 3)),
                        "p = 0.46")),
             hjust = 0, vjust = 1) + 
  theme_bw()
```

### 2c
```{r}
bact_plot = annotate_lm_plot(mod_bacteroidales_lm, "Prenol_Phenylprop", 1.2) + 
  labs(
    x = "Intake of Prenol Lipids,\nPhenylpropanoic Acids\n(mg/1000 kcal)",
    y = expression(atop(italic("Bacteroidales") ~ "Abundance", "Yeo-Johnson Transformed")))
```

Merge and Export Plots
```{r}
combined_plot = ggarrange(gamma_plot, entero_plot, bact_plot, ncol = 3, labels = c("A)", "B)", "C)"), align = "hv")

ggsave(combined_plot, file = 'images/LPS_producers_model_plot.png', 
       units = "in", dpi = 300, width = 9, height = 3.5)
```

