---
title: "B3_alpha_diversity_analyses"
author: "Stephanie Wilson"
date: "April 2025"
output: html_document
---

## SUMMARY
This script prepares data to analyze the relationship between polyphenol intake (total, diversity) and alpha diversity indices (richness and evenness) for FL100 metagenomic data.

## INPUT
  - **FL100_metagenome_Meta.RData** - metadata + metagenomic and polyphenol alpha diversity metrics

## OUTPUT
  - **FL100_alpha_diversity_model_outputs.csv** - Contains summary output from each of the six total models

## DATA PREP
Load Packages
```{r message=FALSE, warning=FALSE}
library(tidyverse); library(ggpubr); library(broom); library(ggResidpanel)
```

Load Data
```{r}
load('RData/FL100_metagenome_Meta.RData')
```

## MICROBRIAL EVENESS
VS TOTAL POLYPHENOL INTAKE
```{r}
evenness_totalintake = lm(mb_evenness ~ pp_daily_mg1000kcal + Age + BMI + Sex+
                         fiber_g1000kcal+hei_asa24_totalscore, data = merge_313)

evenness_totalintake_output = broom::tidy(evenness_totalintake) %>%
    mutate(across(c(estimate:p.value), ~ round(.x, digits =3))) # round

resid_panel(evenness_totalintake)
```

VS POLYPHENOL EVENNESS
```{r}
evenness_pp_evenness= lm(mb_evenness ~ pp_pielous + Age + BMI + Sex+ fiber_g1000kcal+hei_asa24_totalscore, data = merge_313)
summary(evenness_pp_evenness)

# Extract Model Output
evenness_pp_evenness_output = broom::tidy(evenness_pp_evenness) %>%
  mutate(across(c(estimate:p.value), ~ round(.x, digits =3))) # round

resid_panel(evenness_pp_evenness)
```

VS POLYPHENOL DIVERSITY
```{r}
evenness_pp_richness = lm(mb_evenness ~ pp_sobs + Age + BMI + Sex+ fiber_g1000kcal+hei_asa24_totalscore, data = merge_313)
summary(evenness_pp_richness)

# Extract Model Output
evenness_pp_richness_output = broom::tidy(evenness_pp_richness) %>%
  mutate(across(c(estimate:p.value), ~ round(.x, digits =3))) # round

resid_panel(evenness_pp_richness)
```

## MICROBRIAL RICHNESS
VS TOTAL POLYPHENOL INTAKE
```{r}
richness_intake = lm(mb_sobs ~ pp_daily_mg1000kcal + Age + BMI + Sex+ 
                           fiber_g1000kcal + hei_asa24_totalscore, data = merge_313) #%>% filter(pp_daily_mg1000kcal < 4000)

richness_intake_output = broom::tidy(richness_intake) %>%
    mutate(across(c(estimate:p.value), ~ round(.x, digits =3))) # round

resid_panel(richness_intake)
```

VS POLYPHENOL RICHNESS
```{r}
richness_pp_richness = lm(mb_sobs ~ pp_sobs + Age + BMI + Sex+ fiber_g1000kcal + hei_asa24_totalscore, data = merge_313)
  summary(richness_pp_richness)

#Extract Model Output
richness_pp_richness_output = broom::tidy(richness_pp_richness) %>%
  mutate(across(c(estimate:p.value), ~ round(.x, digits =3)))

resid_panel(richness_pp_richness)
```

VS POLYPHENOL EVENNESS
```{r}
richness_pp_evenness = lm(mb_sobs ~ pp_pielous + Age + BMI + Sex+ fiber_g1000kcal+hei_asa24_totalscore, data = merge_313)
summary(richness_pp_evenness)

# Extract Model Output
richness_pp_eveness_output = broom::tidy(richness_pp_evenness) %>%
  mutate(across(c(estimate:p.value), ~ round(.x, digits =3))) # round

resid_panel(richness_pp_evenness)
```


## Save these files for additional plotting if necessary
```{r}
# Create a named list of all dataframes
model_list = list(
  mbevenness_totalintake_output = evenness_totalintake_output,
  mbevenness_pp_evenness_output = evenness_pp_evenness_output,
  mbevenness_pp_richness_output = evenness_pp_richness_output,
  mbrichness_intake_output = richness_intake_output,
  mbrichness_pp_richness_output = richness_pp_richness_output,
  mbrichness_pp_eveness_output = richness_pp_eveness_output
)

# Combine into one dataframe with identifiers
combined_models = bind_rows(model_list, .id = "model_name")

write.csv(combined_models, 'output/FL100_alpha_diversity_model_outputs.csv', row.names = FALSE)
```




